---
title: "R Notebook"
output: html_notebook
---

# library
```{r}
library(tidyverse)
```


# load data
```{r}
tmp1 <- read_csv("merged_df_raw.csv")
```

# filter for only abundance-based studies 
```{r}
tmp1ab <- tmp1 %>%
  select(Biome, Predominant_land_use,
         Use_intensity, Kingdom:Genus,
         Diversity_metric, Diversity_metric_type,
         Effort_corrected_measurement,
         Reference, Site_name, Site_number) %>%
  filter(Diversity_metric_type == "Abundance") %>%
  select(-Diversity_metric, -Diversity_metric_type)
```

# phylogenetic diversity analysis
```{r}
tmp2 <- tmp1ab %>%
  select(Reference, Site_number,
         Kingdom:Genus, Effort_corrected_measurement) %>%
  filter(Effort_corrected_measurement != 0) %>%
  mutate(combined_tax = paste(Kingdom, Phylum,
                              Class, Order,
                              Family, Genus, sep=","))
```

export a list of family names delimited by comma for upload into phyloT
```{r}
exp1 <- tmp2 %>%
  select(Family) %>%
  distinct() 
write.table(exp1$Family, file = "family_names.txt", sep = ",", row.names = FALSE, col.names = FALSE, quote = FALSE)
```

```{r}
unmatched_names <- c("Napoleonaceae", "Carcinophoridae", "Aphodiidae", "Rhipiphoridae", "Oriolidae", 
                  "Melolonthidae", "Cephalotaxaceae", "Cetoniidae", "Troglodytidae", "Emberizidae", 
                  "Anobiidae", "Scydmaenidae", "Coerebidae", "Corytophanidae", "Xanthorrhoeaceae", 
                  "Cracticidae", "Eumenidae", "Aromobatidae", "Agoutidae", "Callitrichidae", 
                  "Neosittidae", "Pangrolaimidae", "Paratropididae", "Viduidae", "Aegithinidae", 
                  "Tropiduridae", "Protoribatidae", "Arctiidae", "Lymantriidae", "Cyphoderidae", 
                  "Melianthaceae", "Dynastidae", "Pteroclididae", "Dryophthoridae", "Opluridae", 
                  "Chrysothricaceae", "Not assigned", "Masaridae", "Austrachipteriidae", "Ptilocercidae", 
                  "Ptilogonatidae", "Languriidae", "Hemiprocnidae", "Bankeraceae", "Phallogastraceae", 
                  "Corethrellidae", "Indridae", "Symmocidae", "Rhyscotidae", "Ascalaphidae", 
                  "Tubiferaceae", "Scoliciosporaceae", "Cinclidotaceae", "Taenitidaceae", "Ganodermataceae", 
                  "Cantacaderidae", "Epimerellidae", "Biatorellaceae", "Eupetidae", "Remizidae", 
                  "Rutelidae", "Thyrisomidae", "Oxydiridae", "Rhynchitidae", "Heterozetidae", 
                  "Stemonitidaceae", "Anomalepidae", "Pudeoniscidae", "Odontorhabditidae", "Labidostommidae", 
                  "Laelaptidae", "Pleuroziopsaceae", "Charipidae", "Paraphelenchidae", "Myoxidae", 
                  "Microzetidae", "Aleurodamaeidae", "Plateremaeidae", "Spinozetidae", "NA")

exp1 <- exp1 %>% 
  filter(!Family %in% unmatched_names) # removing families that are not matched in the database

write.table(exp1$Family, file = "family_names.txt", sep = ",", row.names = FALSE, col.names = FALSE, quote = FALSE) # export to generate phylotree
```

```{r}
# prepare community matrix for derivation of phylogenetic distance scores
community_matrix <- tmp2 %>%
  select(Reference, Site_number, Family, Effort_corrected_measurement) %>%
  mutate(site = paste(Reference, Site_number, sep="_")) %>%
  select(site, Family, Effort_corrected_measurement) %>%
  group_by(site, Family) %>%
  summarize(sum = sum(Effort_corrected_measurement)) %>%
  pivot_wider(names_from = site, values_from = sum) %>%
  t()

colnames(community_matrix) <- community_matrix[1,]
cd <- community_matrix[-1,] %>%
  as.data.frame()
cd <- as.data.frame(apply(cd, 2, as.numeric))

community_matrix_small <- community_matrix[1:100,1:100]  
cd_small <- cd[1:200,c(1:30)]

desired_rownames <- rownames(community_matrix)[2:22755]
rownames(cd) <- desired_rownames
cd <- cd[, !is.na(names(cd)) & names(cd) != ""]
cd <- cd %>% mutate(across(everything(), ~ replace_na(., 0)))
cd <- cd %>% mutate(across(everything(), ~ round(., 0)))
colnames(cd) <- paste0("f__", colnames(cd))

install.packages("ape")
install.packages("picante")l
library(ape); library(picante)
phylo_tree <- read.tree("iphylo_tree.nwk") # import and read tree
family_names <- colnames(cd)
pruned_tree <- drop.tip(phylo_tree, setdiff(phylo_tree$tip.label, family_names)) # drop unmatched families
mismatched_columns <- !(colnames(cd) %in% pruned_tree$tip.label)
mismatched_names <- colnames(cd)[mismatched_columns]
cd <- cd[, !mismatched_columns]

pd_result <- pd(samp =cd, 
                tree = pruned_tree, 
                include.root = TRUE) # generate phylogenetic diversity scores

print(pd_result)
```

```{r}
# export pd scores
write.csv(pd_result, file = "pd_result.csv", row.names = TRUE)
```


# calculating shannon species richness
```{r}
tmp2 <- tmp1ab %>%
  select(Reference, Site_number,Family,
         Genus, Effort_corrected_measurement) %>%
  filter(Effort_corrected_measurement != 0) %>%
  group_by(Reference, Site_number, Family, Genus) %>%
  summarize(sum= sum(Effort_corrected_measurement)) %>%
  mutate(sum = round(sum,0)) %>%
  ungroup() %>%
  group_by(Reference, Site_number) %>%
  mutate(site_total = sum(sum)) %>%
  ungroup() %>%
  mutate(taxon_prop = sum/site_total) %>%
  mutate(ln_taxon_prop = log(taxon_prop)) %>%
  mutate(prop_lnprop = taxon_prop*ln_taxon_prop) %>%
  group_by(Reference, Site_number) %>%
  select(Reference, Site_number, prop_lnprop) %>%
  summarize(shannon_rich = sum(prop_lnprop)*-1) %>%
  mutate(ref_site = paste(Reference, Site_number, sep="_"))%>%
  ungroup() %>%
  select(ref_site, shannon_rich)

shann_rich <- tmp2
```

## extra check:  number of studies that form abundance

```{r}
tmp <- tmp1 %>%
  select(Reference, Site_number, Diversity_metric_type) %>%
  distinct()

table1 <- table(tmp$Diversity_metric_type)
print(table1) # by sites
```

```{r}
tmp <- tmp1 %>%
  select(Reference, Diversity_metric_type) %>%
  distinct()

table1 <- table(tmp$Diversity_metric_type)
print(table1) # by sites
```

# calculating habitat quality

```{r}
tmp2 <- tmp1ab %>%
  select(Reference, Site_number,
         Biome, Predominant_land_use,
         Use_intensity) %>%
  mutate(study_site = paste(Reference, Site_number, sep = "_")) %>%
  distinct() %>%
  select(study_site, Biome, Predominant_land_use, Use_intensity) %>%
  mutate(lu_score = 0) %>%
  filter(Predominant_land_use != "Cannot decide")
  
# assign scores based on current land use
if (ncol(tmp2) >= 5) {
  condition <- tmp2[, 2] %in% c("Mature secondary vegetation",
                                "Secondary vegetation (indeterminate age)",
                                "Intermediate secondary vegetation",
                                "Primary vegetation",
                                "Young secondary vegetation")
    tmp2[, 5] <- ifelse(condition, 3, 1)
} else {
  warning("df does not have correct columns")
}

tmp2 <- tmp2 %>%
  filter(Use_intensity != "Cannot decide") %>%
  mutate(ui_score = 0)

# assign scores based on use intensity
for(i in 1:nrow(tmp2)){
  if(tmp2[i,4] == "Minimal use"){
    tmp2[i,6] = -0.1
  }
  if(tmp2[i,4] == "Light use"){
    tmp2[i,6] = -0.3
  }
  if(tmp2[i,4] == "Intense use"){
    tmp2[i,6] = -0.5
  }
}

tmp2 <- tmp2 %>%
  mutate(hq_score = lu_score + ui_score)

hqi_score = tmp2 %>%
  select(study_site, hq_score)
```

# HANPP

```{r}
tmp2 <- tmp1ab %>%
  select(Reference, Site_number,
         Biome, Predominant_land_use, Use_intensity) %>%
  mutate(ref_site = paste(Reference, Site_number, sep="_")) %>%
  select(-Reference, -Site_number) %>%
  distinct() %>%
  select(ref_site, Biome, Predominant_land_use, 
         Use_intensity) %>%
  mutate(bl_prod = "TBC",
         app_prod = "TBC")
  
condition <- tmp2$Biome %in% c(
  "Tropical & Subtropical Moist Broadleaf Forests",
  "Tropical & Subtropical Grasslands, Savannas & Shrublands",
  "Tropical & Subtropical Dry Broadleaf Forests",
  "Tropical & Subtropical Coniferous Forests",
  "Mangroves")
  tmp2$bl_prod <- ifelse(condition, "high", "TBC")
    
condition <- tmp2$Biome %in% c(
  "Temperate Broadleaf & Mixed Forests",
  "Mediterranean Forests, Woodlands & Scrub",
  "Temperate Conifer Forests",
  "Temperate Grasslands, Savannas & Shrublands",
  "Flooded Grasslands & Savannas")
  tmp2$bl_prod <- ifelse(condition, "medium", tmp2$bl_prod)
    
condition <- tmp2$Biome %in% c(
    "Montane Grasslands & Shrublands",
    "Deserts & Xeric Shrublands",
    "Boreal Forests/Taiga",
    "Tundra")
    tmp2$bl_prod <- ifelse(condition, "low", tmp2$bl_prod)
    
# account for use_intensity
app_prod_condition <- tmp2$Predominant_land_use %in% c(
  "Mature secondary vegetation",
  "Secondary vegetation (indeterminate age)",
  "Intermediate secondary vegetation",
  "Primary vegetation",
  "Young secondary vegetation")
    tmp2$app_prod <- ifelse(app_prod_condition, "low", "TBC")

app_prod_condition <- tmp2$Predominant_land_use %in% c("Urban")    
  
#assign HANPP category based on land use and intensity
tmp2$app_prod <- ifelse(app_prod_condition, "high", tmp2$app_prod)  
  
  app_prod_condition <- c(tmp2$Predominant_land_use %in% c("Cropland",
                                                         "Plantation forest",
                                                         "Pasture") & tmp2$Use_intensity %in% c("Minimal use",
                                                                                                "Light use"))    
  tmp2$app_prod <- ifelse(app_prod_condition, "low", tmp2$app_prod)  
  
app_prod_condition <- c(tmp2$Predominant_land_use %in% c("Cropland",
                                                         "Plantation forest",
                                                         "Pasture") & tmp2$Use_intensity %in% c("Intense use"))    
  tmp2$app_prod <- ifelse(app_prod_condition, "high", tmp2$app_prod)  

#just to see number of sites without land use    
tmp3 <- tmp2 %>%
  filter(Predominant_land_use == "Cannot decide") %>%
  select(ref_site) %>%
  distinct()


hanpp_score <- tmp2 %>%
  select(ref_site, app_prod)
```


# PDF
by site instead, and columns will be classes present
```{r}
tmp2 <- tmp1 %>%
  select(Reference, Site_number,
         Kingdom, Phylum, Class, Order, Family, Genus,
         Biome, Predominant_land_use,
         Order, Effort_corrected_measurement,
         Diversity_metric_type) %>%
  mutate(ref_site = paste(Reference, Site_number, sep= "_"),
         conversion = "undefined") %>%
  filter(Predominant_land_use != "Cannot decide",
         Diversity_metric_type == "Abundance")

# projecting loss at class level
tmp2b <- tmp2 %>%
  select(ref_site, Class, Biome, Predominant_land_use, 
         Effort_corrected_measurement)

tmp2c <- tmp2b %>%
  group_by(ref_site, Class, Biome, Predominant_land_use) %>%
  summarize(total = sum(Effort_corrected_measurement)) %>%
  mutate(total = round(total, 2),
         lu = Predominant_land_use,
         status = "pending") %>%
  ungroup()


if (ncol(tmp2c) >= 4) {
  # Initialize the 6th column with the existing values from the 5th column
  tmp2c[, 7] <- tmp2c[, 6]
  
  for (i in 1:nrow(tmp2c)) {
    if (tmp2c[i, 6] == "Primary vegetation") {
      tmp2c[i, 7] <- "primary baseline"
    } else if (tmp2c[i, 6] %in% c("Young secondary vegetation",
                                  "Intermediate secondary vegetation",
                                  "Mature secondary vegetation",
                                  "Secondary vegetation (indeterminate age)")) {
      tmp2c[i, 7] <- "secondary baseline"
    }
  }
}

tmp2d <- tmp2c %>%
  select(ref_site, Class, Biome, status, total) %>%
  group_by(ref_site, Class, Biome, status) %>%
  summarize(mean = mean(total, na.rm = T)) %>%
  mutate(mean = round(mean, 0)) %>%
# summarize for secondary baseline
  group_by(ref_site, Class, Biome, status) %>%
  pivot_wider(names_from = status, values_from = mean) %>%
  # pivot out baselines to deduce change in values based on LUC
  filter(!(is.na(`primary baseline`) & is.na(`secondary baseline`))) %>%
  rename(`primary_b` = `primary baseline`,
         `secondary_b` = `secondary baseline`) %>%
  mutate(cropland_pri = (Cropland - primary_b)/(primary_b+0.0000001)*100, # adding 0.0000001 to avoid division by 0 leading to inf values
         pasture_pri = (Pasture - primary_b)/(primary_b+0.0000001)*100,
         plantation_pri = (`Plantation forest` - primary_b)/(primary_b+0.0000001)*100,
         urban_pri = (Urban - primary_b)/(primary_b+0.0000001)*100,
         cropland_sec = (Cropland - secondary_b)/(secondary_b+0.0000001)*100,
         plantation_sec = (`Plantation forest` - secondary_b)/(secondary_b+0.0000001)*100,
         pasture_sec = (Pasture - secondary_b)/(secondary_b+0.0000001)*100,
         urban_sec = (Urban - secondary_b)/(secondary_b+0.0000001)*100,
) %>%
    mutate(across(c(cropland_pri, 
                    pasture_pri,
                    plantation_pri,
                    urban_pri,
                    cropland_sec,
                    plantation_sec,
                    pasture_sec,
                    urban_sec), round, 0))

tmp2d2 <- tmp2d

# summing the changes and taking the mean for reduced dimensions
tmp2d2$mean_avg<- rowMeans(tmp2d2[, (ncol(tmp2d2)-7):ncol(tmp2d2)], na.rm = T)

# replacing NaN values with 0
tmp2d2$mean_avg <- ifelse(is.nan(tmp2d2$mean_avg), 0, tmp2d2$mean_avg)

tmp2d2 <- tmp2d %>%
  group_by(ref_site, Biome) %>%
  select(ref_site, Biome) %>%
  distinct()

# it seems there are sites that have more than one biome - let's take a look at them
tmp2d2$ref_site <- trimws(tolower(tmp2d2$ref_site))

duplicated_rows <- tmp2d2[duplicated(tmp2d2$ref_site) | duplicated(tmp2d2$ref_site, fromLast = TRUE), ]

# let's identify these sites 
unwanted_sites <- unique(duplicated_rows$ref_site)

tmp2e <- tmp2d2[,c(1:3, 18)] %>%
  ungroup() %>%
  filter(is.na(Class) != TRUE) %>%
  select(-Biome) %>%
  filter(!(ref_site %in% unwanted_sites)) # removing these sites
  
  # the final step is to join the sites with the columns of abd change back to combined_scoresv2

# continued  below at combined_scoresv2 area
  
```


## Old PDF - by Class (above and below 80% loss in abundance)

```{r}
tmp2 <- tmp1 %>%
  select(Reference, Site_number,
         Kingdom, Phylum, Class, Order, Family, Genus,
         Biome, Predominant_land_use,
         Order, Effort_corrected_measurement,
         Diversity_metric_type) %>%
  mutate(Study_site = paste(Reference, Site_number, sep= "_"),
         conversion = "undefined") %>%
  filter(Predominant_land_use != "Cannot decide",
         Diversity_metric_type == "Abundance")

# projecting loss at class
tmp2b <- tmp2 %>%
  select(Class, Biome, Predominant_land_use, 
         Effort_corrected_measurement) %>%
  group_by(Class, Biome, Predominant_land_use) %>%
  summarize(total = sum(Effort_corrected_measurement)) %>%
  mutate(total = round(total, 2),
         lu = Predominant_land_use,
         status = "pending") %>%
  ungroup()

if (ncol(tmp2b) >= 4) {
  # Initialize the 6th column with the existing values from the 5th column
  tmp2b[, 6] <- tmp2b[, 5]
  
  for (i in 1:nrow(tmp2b)) {
    if (tmp2b[i, 5] == "Primary vegetation") {
      tmp2b[i, 6] <- "primary baseline"
    } else if (tmp2b[i, 5] %in% c("Young secondary vegetation",
                                  "Intermediate secondary vegetation",
                                  "Mature secondary vegetation",
                                  "Secondary vegetation (indeterminate age)")) {
      tmp2b[i, 6] <- "secondary baseline"
    }
  }
}

tmp2c <- tmp2b %>%
  select(Class, Biome, status, total) %>%
  group_by(Class, Biome, status) %>%
  summarize(mean = mean(total, na.rm = T)) %>%
  mutate(mean = round(mean, 0)) %>%
# summarize for secondary baseline
  group_by(Class, Biome, status) %>%
  pivot_wider(names_from = status, values_from = mean) %>%
  # pivot out baselines to deduce change in values based on LUC
  filter(!(is.na(`primary baseline`) & is.na(`secondary baseline`))) %>%
  rename(`primary_b` = `primary baseline`,
         `secondary_b` = `secondary baseline`) %>%
  mutate(cropland_pri = (Cropland - primary_b)/primary_b*100,
         pasture_pri = (Pasture - primary_b)/primary_b*100,
         plantation_pri = (`Plantation forest` - primary_b)/primary_b*100,
         urban_pri = (Urban - primary_b)/primary_b*100,
         cropland_sec = (Cropland - secondary_b)/secondary_b*100,
         plantation_sec = (`Plantation forest` - secondary_b)/secondary_b*100,
         pasture_sec = (Pasture - secondary_b)/secondary_b*100,
         urban_sec = (Urban - secondary_b)/secondary_b*100,
) %>%
    mutate(across(c(cropland_pri, 
                    pasture_pri,
                    plantation_pri,
                    urban_pri,
                    cropland_sec,
                    plantation_sec,
                    pasture_sec,
                    urban_sec), round, 0))


tmp2c <- tmp2c[,c(1, 2, 9, 10, 11, 12, 13, 14, 15, 16)]
print(tmp2c)
# deduce percentages

# BUT WE NEED SET THRESHOLD - how many species lost by 80-90%
```

first we split into primary and secondary baselines

```{r}
tmp2c_pri <- tmp2c %>%
  select(Class:urban_pri)

tmp2c_sec <- tmp2c %>%
  select(Class, Biome, cropland_sec:urban_sec)
```

then we break those into above 80 and below 80%
  first we pivot
  
```{r}
# primary baselines
# below 80
tmp2c_pri_b80 <- tmp2c_pri %>%
  pivot_longer(cols=c("cropland_pri",
               "pasture_pri",
               "plantation_pri",
               "urban_pri"),
               names_to = "conversion",
               values_to = "pct",
               values_drop_na = T) %>%
  filter(!(pct == Inf),
         pct <= -80)

# above 80
tmp2c_pri_a80 <- tmp2c_pri %>%
  pivot_longer(cols=c("cropland_pri",
               "pasture_pri",
               "plantation_pri",
               "urban_pri"),
               names_to = "conversion",
               values_to = "pct",
               values_drop_na = T) %>%
  filter(!(pct == Inf),
         pct > -80) %>%
  arrange(desc(pct))
```

```{r}
# secondary baselines
# below 80
tmp2c_sec_b80 <- tmp2c_sec %>%
  pivot_longer(cols=c("cropland_sec",
               "pasture_sec",
               "plantation_sec",
               "urban_sec"),
               names_to = "conversion",
               values_to = "pct",
               values_drop_na = T) %>%
  filter(!(pct == Inf),
         pct <= -80)

# above 80
tmp2c_sec_a80 <- tmp2c_sec %>%
  pivot_longer(cols=c("cropland_sec",
               "pasture_sec",
               "plantation_sec",
               "urban_sec"),
               names_to = "conversion",
               values_to = "pct",
               values_drop_na = T) %>%
  filter(!(pct == Inf),
         pct > -80) %>%
  arrange(desc(pct))
```

the taxonomic groups with outlier values are reasonable (plants, conifer pines in plantations, nematodes, insects and smaller organisms like archanids which often have larger populations and larger magnitude of change)


### extracting classes

```{r}
tmp2c_pri_neg <- tmp2c_pri %>%
  pivot_longer(cols=c("cropland_pri",
               "pasture_pri",
               "plantation_pri",
               "urban_pri"),
               names_to = "conversion",
               values_to = "pct",
               values_drop_na = T) %>%
  filter(!(pct == Inf),
         pct < 0) %>%
  arrange(pct) %>%
  filter(!is.na(Class), Class != "Not assigned")

tmp2c_sec_neg = tmp2c_sec %>%
  pivot_longer(cols=c("cropland_sec",
               "pasture_sec",
               "plantation_sec",
               "urban_sec"),
               names_to = "conversion",
               values_to = "pct",
               values_drop_na = T) %>%
  filter(!(pct == Inf),
         pct < 0)%>%
  arrange(pct) %>%
  filter(!is.na(Class), Class != "Not assigned")
```

# Mean Spp Abundance
total number of individuals of each species and dividing it by the number of species present.

at site level

```{r}
tmp2 <- tmp1ab %>%
  select(Reference, Site_number, 
         Family, Genus, 
         Biome, Predominant_land_use,
         Effort_corrected_measurement) %>%
  filter(!(Effort_corrected_measurement == 0)) %>%
  mutate(ref_site = paste(Reference, Site_number, sep = "_"),
         fam_genus = paste(Family, Genus, sep = "_")) %>%
  select(-Reference, -Site_number, -Family, -Genus) %>%
  group_by(ref_site,
           Biome,
           Predominant_land_use,
           fam_genus) %>%
  summarize(total = sum(Effort_corrected_measurement)) %>%
  group_by(ref_site, Biome, Predominant_land_use, fam_genus) %>%
  summarize(site_total = round(sum(total),0)) %>%
   group_by(ref_site) %>%
  mutate(site_total_spp = n_distinct(fam_genus)) %>%


  group_by(ref_site, Biome, Predominant_land_use) %>%
  summarize(site_total_ab = sum(site_total),
             site_total_spp = first(site_total_spp))
  
# calculate mean species abundance
tmp3 <- tmp2 %>%
  select(ref_site,
         site_total_ab,
         site_total_spp) %>%
  ungroup() %>%
  select(-Biome) %>%
  mutate(msa = round(site_total_ab/site_total_spp,0)) %>%
  select(ref_site, msa)

msa <- tmp3
```

# data integration

```{r}
pd_result <- pd_result %>%
  rownames_to_column(var="ref_site") %>%
  select(ref_site, PD)

hqi_score <- hqi_score %>%
  mutate(ref_site = study_site) %>%
  select(ref_site, hq_score)

hqi_score <- hqi_score %>%
  group_by(ref_site) %>%
  slice_min(hq_score)

combined_scores <- hanpp_score %>%
  inner_join(., msa) %>%
  inner_join(., hqi_score) %>%
  inner_join(., shann_rich) %>%
  inner_join(., pd_result) %>%
  inner_join(., tmp2e)

combined_scores <- combined_scores %>%
  rename(hanpp = app_prod,
         spp_rich = shannon_rich,
         phylo_div = PD)

write.csv(combined_scores, "combined_scoresv1.csv")
```

# addition of coordinates
```{r}
tmp3 <- tmp1 %>%
  select(Reference, Site_number, Longitude, Latitude) %>%
  distinct() %>%
  mutate(ref_site = paste(Reference, Site_number, sep="_")) %>%
  select(ref_site, Latitude, Longitude)

combined_scores <- combined_scores %>%
  left_join(., tmp3)

write.csv(combined_scores, "combined_scoresv1.csv")
```


## filtering out sites with multiple coordinates
```{r}
coord <- tmp1 %>%
  select(Latitude, Longitude, Reference, Site_number) %>%
  distinct()

coord2 <- tmp1 %>%
  select(Latitude, Longitude, Reference, Site_number) %>%
  mutate(ref_site = paste(Reference, Site_number, sep="_")) %>%
  select(ref_site, Latitude, Longitude) %>%
  distinct() %>%
  group_by(ref_site) %>%
  summarize(n = n())

table((coord2$n))/26699*100
```
if we only retain the sites with only 1 pair of coordinates, we will still have 91% of sites

so lets filter out the sites with more than 1 set of spatial coordinates
```{r}
coord3 = coord2 %>%
  filter( n == 1)

combined_scores2 <- combined_scores %>%
  filter(ref_site %in% coord3$ref_site)

rm(coord, coord2, coord3)
```


```{r}
write.csv(combined_scores2, "combined_scoresv2.csv")
```

adding in pdf
```{r}
combined_scores3 <- combined_scores2
combined_scores3$ref_site <- trimws(tolower(combined_scores3$ref_site))
tmp2e <- tmp2e %>%
  select(ref_site, mean_avg)
combined_scores3 <- combined_scores3 %>%
  #left_join(., tmp2e) %>%
  filter(hanpp != "TBC") %>% # remove undecided hanpp sites
  filter(!is.nan(spp_rich)) # removing residue sites with NaN spp richness

# combined_scores3$mean_avg = ifelse(is.na(combined_scores3$mean_avg), 0, combined_scores3$mean_avg) # some pdf has na, replace with 0
```

```{r}
tmp2e_non0 <- tmp2e %>%
  filter(mean_avg != 0)
```

```{r}
table(tmp2e_non0$ref_site %in% combined_scores$ref_site)
```
So it seems non of the sites that recorded changes in abundance of organisms across a change in land use are in the combined_scores

and attempts to inner_join led revealed that both df ref_sites are mutually_exclusive. we are talking about only 6 studies or so that actually did the measurement across different luc

# Modeling using multiple linear reg

```{r}
str(combined_scores3)
```

```{r}
combined_scores4 = combined_scores3 %>%
  mutate(hanpp = as.factor(hanpp))

combined_scores4 <- cbind(combined_scores4, model.matrix(~ hanpp - 1, data = combined_scores4)) # convert matrix with onehot encoding for hanpp scores (high or low)

combined_scores4$log_msa = log(combined_scores4$msa + 1) # applying log transformation to deal with outliers of msa

## scaling

### defining variables
x- xmin/(xmax-xmin)
spp_rich_min = min(combined_scores4$spp_rich, na.rm = T)
spp_rich_max = max(combined_scores4$spp_rich, na.rm = T)
phylo_div_min = min(combined_scores4$phylo_div, na.rm = T)
phylo_div_max = max(combined_scores4$phylo_div, na.rm = T)
log_msa_min = min(combined_scores4$log_msa, na.rm = T)
log_msa_max = max(combined_scores4$log_msa, na.rm = T)

combined_scores4 =  combined_scores4 %>%
  mutate(
    spp_rich_scaled = (spp_rich - spp_rich_min)/(spp_rich_max - spp_rich_min),
    phylo_div_scaled = (phylo_div - phylo_div_min)/(phylo_div_max - phylo_div_min),
    log_msa_scaled = (log_msa - log_msa_min)/(log_msa_max - log_msa_min),
    ov_score = hq_score + spp_rich_scaled + phylo_div_scaled + log_msa_scaled + hanpphigh*0.2 + hanpplow*0.8) 

# let's proceed to put in a multiple regression model

model <- lm(ov_score ~ hq_score + spp_rich_scaled + phylo_div_scaled + 
               + log_msa_scaled + hanpplow + hanpphigh, 
            data = combined_scores4)

# View the model summary
summary(model)
```



















