---
title: "Database Filtering and Taxonomic Classification"
author: "Eli Duggan"
date: "2025-06-14"
output: html_notebook
---

# This script should be run second.

# 1. Load packages
```{r setup, include=FALSE}
pkgs <- c("stringdist", "taxize", "vroom", "data.table", "furrr")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)

library(stringdist)
library(taxize)
library(vroom)
library(data.table)
library(furrr)
```

# 2. Load data
```{r}
# Load BioTIME data
data <- readRDS("../data/biotime_v2_terrestrial_post_2014.rds")
data <- as.data.table(data)

# Load GBIF Taxon.tsv for GBIF species list - UPDATE
gbif_file <- "../data/Taxon.tsv"
gbif_data <- fread(gbif_file, quote = "")

# Filter & build full_binomial fast - UPDATE
gbif_species <- gbif_data[
  taxonRank == "species" & !is.na(genus) & !is.na(specificEpithet),
  .(full_binomial = paste(genus, specificEpithet))]

# Deduplicate immediately
setkey(gbif_species, full_binomial)
gbif_species <- unique(gbif_species)
```

# 3. Define helper functions

### Filter unusable data (data.table function)
```{r}
filter_species <- function(dt, species_column = "valid_name") {

  original_count <- nrow(dt)
  
  dt_filtered <- dt

  # Build expression dynamically for the species column
  species_vec <- dt[[species_column]]

  invalid_terms <- c(
  "morphospecies", "unknown", "undetermined", "unidentif",
  "sp\\.", "spp\\.", "sp$", "cf\\.", "indet\\.", "sp[0-9a-zA-Z]+")
  pattern <-paste(invalid_terms, collapse = "|")

  dt_filtered <- dt[
  !is.na(species_vec) &
  !grepl(pattern, species_vec, ignore.case = TRUE) &
  (is.na(resolution) |
    !(tolower(resolution) %chin% c("phylum", "subphylum", "superclass", "kingdom")))]

  kept_count <- nrow(dt_filtered)
  removed_count <- original_count - kept_count
  removed_pct <- round((removed_count / original_count) * 100, 2)

  cat("Original rows:", original_count, "\n")
  cat("Rows removed:", removed_count, "\n")
  cat("Rows kept:", kept_count, "\n")
  cat("Percentage removed:", removed_pct, "%\n")

  return(dt_filtered)
}
```

### Fast bird names resolution function
```{r}
resolve_bird_names <- function(dt, species_column = "valid_name", bird_lookup_path = "../input_files/bird_codes.rds") {

  # Load bird lookup from .rds instead of CSV
  bird <- readRDS(bird_lookup_path)

  # Prep the lookup table
  bird[, `:=`(
    four_code = toupper(trimws(SPEC)),
    six_code = toupper(trimws(SPEC6)),
    common_clean = tolower(gsub("[^a-z]", "", COMMONNAME)),
    scientific_name = trimws(SCINAME)
  )]

  dt[, input_upper := toupper(trimws(get(species_column)))]
  dt[, input_clean := tolower(gsub("[^a-z]", "", get(species_column)))]

  # 4-letter code
  dt[is.na(resolution_method) & input_upper %in% bird$four_code, 
     `:=`(resolved_name = bird[.SD, on = .(four_code = input_upper), x.scientific_name],
          resolution_method = "bird_4letter")]

  # 6-letter code
  dt[is.na(resolution_method) & input_upper %in% bird$six_code, 
     `:=`(resolved_name = bird[.SD, on = .(six_code = input_upper), x.scientific_name],
          resolution_method = "bird_6letter")]

  # Common name
  dt[is.na(resolution_method) & input_clean %in% bird$common_clean, 
     `:=`(resolved_name = bird[.SD, on = .(common_clean = input_clean), x.scientific_name],
          resolution_method = "bird_common")]

  dt[, c("input_upper", "input_clean") := NULL]
}
```

### Fast plant code resolution function
```{r}
decode_plant_codes <- function(dt, species_column = "valid_name", plant_lookup_path = "../input_files/plant_codes.rds") {

  plant_lookup <- readRDS(plant_lookup_path)
  plant_lookup <- plant_lookup[!is.na(code8) & code8 != ""]
  plant_lookup <- plant_lookup[!is.na(plant_binomial) & plant_binomial != ""]
  
  plant_lookup[, code8 := toupper(code8)]
  plant_lookup[, code7_3_4 := toupper(code7_3_4)]
  plant_lookup[, code7_4_3 := toupper(code7_4_3)]

  dt[, code := ifelse(nchar(get(species_column)) %in% c(7, 8) & grepl("^[A-Z]+$", get(species_column)),
                      get(species_column), NA_character_)]

  # Exact 8-char match
  dt[is.na(resolution_method) & code %in% plant_lookup$code8, 
     `:=`(resolved_name = plant_lookup[.SD, on = .(code8 = code), x.plant_binomial],
          resolution_method = "plant_8char")]

  # Exact 7-char 3+4 match
  dt[is.na(resolution_method) & code %in% plant_lookup$code7_3_4, 
     `:=`(resolved_name = plant_lookup[.SD, on = .(code7_3_4 = code), x.plant_binomial],
          resolution_method = "plant_7char")]

  # Exact 7-char 4+3 match
  dt[is.na(resolution_method) & code %in% plant_lookup$code7_4_3, 
     `:=`(resolved_name = plant_lookup[.SD, on = .(code7_4_3 = code), x.plant_binomial],
          resolution_method = "plant_7char")]

  # Fuzzy (optional, leave this for now)
  dt[, code := NULL]
}
```

### Fast spellcheck function w/fuzzy matching
```{r}
spellcheck_species_names <- function(dt, species_column = "valid_name", max_dist = 2) {

  set.seed(8)

  unresolved_names <- unique(na.omit(dt[is.na(resolution_method), get(species_column)]))
  if (length(unresolved_names) == 0) return()

  plan(multisession, workers = parallel::detectCores() - 1)
  col_results <- future_map_dfr(unresolved_names, function(name) {
    tryCatch({
      res <- taxize::col_search(name, limit = 1)
      if (nrow(res) > 0) {
        data.frame(input_name = name, col_canonical = res$name, col_id = res$id, stringsAsFactors = FALSE)} 
      else {
        data.frame(input_name = name, col_canonical = NA_character_, col_id = NA, stringsAsFactors = FALSE)}
    }, error = function(e) {
      data.frame(input_name = name, col_canonical = NA_character_, col_id = NA, stringsAsFactors = FALSE)})
  })

  resolved <- col_results[!is.na(col_results$col_canonical), ]

  lookup <- as.data.table(resolved[, c("input_name", "col_canonical")])
  setnames(lookup, "input_name", species_column)
  setnames(lookup, "col_canonical", "final_name")
  setkeyv(lookup, species_column)
  setkeyv(dt, species_column)

  dt[lookup, on = species_column, 
     `:=`(resolved_name = i.final_name, resolution_method = "spellcheck")]
}
```


### Master resolution function
```{r}
resolve_species_names <- function(dt, 
                                   species_column = "valid_name", 
                                   bird_lookup_path = "../input_files/bird_codes.rds",
                                   plant_lookup_path = "../input_files/plant_codes.rds",
                                   manual_file_path = NULL) {
  
  cat("Starting species name resolution pipeline.\n")

  cat("Resolving bird names...\n")
  resolve_bird_names(dt, species_column, bird_lookup_path)
  cat("Resolving plant names...\n")
  decode_plant_codes(dt, species_column, plant_lookup_path)
  cat("Resolving misspelled names...\n")
  spellcheck_species_names(dt, species_column)

  cat("Species name resolution complete.\n")
}
```

# 4. Filtering and species resolution pipeline
```{r}
# filtered_data <- filter_species(biotime_data_subset) # testing
filtered_data <- filter_species(biotime_data)
```

# Initialize columns and tag already valid binomials
```{r}
# Initialize resolved_name as empty, and pre-tag already valid binomials
filtered_data[, `:=`(
  resolved_name = NA_character_,
  resolution_method = NA_character_)]

# Then, pre-tag already valid GBIF binomials only where we have a confident match
filtered_data[valid_name %chin% gbif_species$full_binomial, `:=`(
  resolved_name = valid_name,
  resolution_method = "already_binomial")]

filtered_data[valid_name %chin% gbif_species$full_binomial, .N]
```

# Resolve species names using helper functions
```{r}
resolve_species_names(filtered_data, species_column = "valid_name")
```

# 5. Diagnostics
```{r}
species_resolution_diagnostics <- function(dt) {

  total_rows <- nrow(dt)
  unresolved_rows <- dt[is.na(resolution_method), .N]
  
  # Build method counts table
  method_counts <- dt[, .N, by = resolution_method][order(-N)]
  
  # Replace NA with "Unresolved" for display
  method_counts[is.na(resolution_method), resolution_method := "Unresolved"]
  
  # Add percentages
  method_counts[, Percent := sprintf("%.1f%%", 100 * N / total_rows)]
  
  cat("\n--- Species Name Resolution Summary ---\n")
  cat("Total rows:      ", total_rows, "\n")
  cat("Unresolved rows: ", unresolved_rows, "\n\n")
  
  print(method_counts, row.names = FALSE)
  
  if (unresolved_rows > 0) {
  cat("\nUnresolved entries:\n")
  print(dt[is.na(resolution_method), .(valid_name)], row.names = FALSE)}
}

species_resolution_diagnostics(filtered_data)
```

# 6. Export outputs

# note - may need to adjust exports to include not just valid/resolved name and how resolved, but also the rest of the database information that is retained.
```{r}
# Convert data to data.table
dt <- as.data.table(filtered_data)

# 1. Export unresolved names
unresolved <- dt[is.na(resolution_method), unique(valid_name)]
saveRDS(data.table(valid_name = unresolved), "../data/unresolved_names.rds")

# 2. Export resolved names
resolved <- unique(dt[!is.na(resolution_method), .(valid_name, resolved_name, resolution_method)])
saveRDS(resolved, "../data/resolved_names.rds")

# 3. Extract resolved names for taxonomic classification
dt_binomials <- dt[!is.na(resolved_name) & grepl("^\\S+\\s+\\S+", resolved_name)]

# Extract genus and species
dt_binomials[, c("genus", "species") := tstrsplit(resolved_name, "\\s+", keep = 1:2)]
dt_binomials[, genus_species := paste(genus, species)]

# Deduplicate unique genus_species
dt_unique <- unique(dt_binomials, by = "genus_species")

# Assign row numbers and reorder
setorder(dt_unique, genus, species)
dt_unique[, number := .I]
result <- dt_unique[, .(number, genus, species, genus_species)]

# Export unique genus_species for taxonomy step
saveRDS(result, "../data/unique_species.rds")

```