---
title: "biodiversity_defor_tagging"
author: "Eli Duggan"
date: "2026-03-04"
output: html_document
---

## Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
library(sf)
library(tidyverse)
library(maps)
```

## Configuration
```{r}

# Directories
in_dir <- file.path("..", "input")
out_dir <- file.path("..", "output")
tmp_dir <- file.path("..", "tmp")
defor_spatial_dir <- file.path("..", "..", "spatial_data", "deforestation", "spatial")
defor_tabular_dir <- file.path("..", "..", "spatial_data", "deforestation", "tabular")
cluster_dir <- file.path("..", "..", "clustering", "output")

dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)


# Analysis settings
buffer_km_vals <- c(1, 3, 5, 10)
analysis_crs <- 6933


# Clustering input selection
cluster_method <- "clara"
cluster_radius <- "12.5km"
cluster_file <- paste0(cluster_method, "_rad_", cluster_radius, "_model_df_clustered.csv")


# Input files
defor_spatial_file <- file.path(defor_spatial_dir, "deforestation_50km.shp")
defor_tabular_file <- file.path(defor_tabular_dir,"deforestation_tile_total_ha_SPAM2020_2001_2024.csv")
cluster_file_path <- file.path(cluster_dir, cluster_file)

# Output files
gpkg_path <- file.path(out_dir, "biodiversity_deforestation_layers.gpkg")


# Lightweight caches for expensive derived objects only
defor_tiles_cache <- file.path(tmp_dir, "defor_tiles_sf.rds")
cluster_footprints_cache <- file.path(tmp_dir, "cluster_footprints_sf.rds")

read_or_build <- function(path, build_fn, rebuild = FALSE) {
  if (!rebuild && file.exists(path)) {
    message("Loading cached object: ", basename(path))
    return(readRDS(path))
  }

  message("Building object: ", basename(path))
  value <- build_fn()
  saveRDS(value, path)
  value
}
```

## Load input data
```{r}
#-----------------------------
# Small validation helpers
#-----------------------------
assert_has_cols <- function(data, cols, data_name) {
  missing_cols <- setdiff(cols, names(data))
  if (length(missing_cols) > 0) {
    stop(
      paste0(
        data_name, " is missing required columns: ",
        paste(missing_cols, collapse = ", ")),
      call. = FALSE)}}

assert_no_duplicate_keys <- function(data, keys, data_name) {
  duplicate_rows <- data %>%
    count(across(all_of(keys)), name = "n") %>%
    filter(n > 1)

  if (nrow(duplicate_rows) > 0) {
    stop(
      paste0(
        data_name, " has duplicate key rows for: ",
        paste(keys, collapse = ", ")
      ),
      call. = FALSE)}}


#-----------------------------
# Read raw inputs
#-----------------------------
defor_spatial_raw <- read_sf(defor_spatial_file) %>% 
  st_make_valid() %>% 
  mutate(id = trimws(as.character(id)))

defor_tabular_raw <- read_csv(defor_tabular_file, show_col_types = FALSE)

cluster_raw <- read_csv(cluster_file_path, show_col_types = FALSE)


#-----------------------------
# Check required columns
#----------------------------
assert_has_cols(
  defor_spatial_raw, 
  c("id", "geometry"),
  "defor_spatial_raw"
  )

assert_has_cols(
  defor_tabular_raw,
  c("id", "lossyear_mode", "total_deforested_ha"),
  "defor_tabular_raw"
)

assert_has_cols(
  cluster_raw,
  c("sample_id", "AEZ", "cluster_id", "Latitude", "Longitude", "year", "ov_score"),
  "cluster_raw"
)

colnames(defor_spatial_raw)
colnames(defor_tabular_raw)
colnames(cluster_raw)


#-----------------------------
# Build canonical tile table
# one row per tile
#-----------------------------

tiles_with_ha <- defor_tabular_raw %>%
  transmute(
    tile_id = as.character(id),
    has_ha_info = !is.na(total_deforested_ha)
  ) %>%
  group_by(tile_id) %>%
  summarise(
    has_ha_info = any(has_ha_info),
    .groups = "drop"
  )

defor_tiles <- defor_spatial_raw %>%
  transmute(
    tile_id = as.character(id),
    iso3,
    country_name = name,
    region,
    admin1,
    adm1_name,
    eco_name = ECO_NAME,
    biome_name = BIOME_NAME,
    realm = REALM,
    tile_area_raw = area,
    geometry
  ) %>%
  left_join(tiles_with_ha, by = "tile_id") %>%
  mutate(
    has_ha_info = coalesce(has_ha_info, FALSE)
  ) %>%
  st_transform(analysis_crs)

assert_no_duplicate_keys(
  st_drop_geometry(defor_tiles),
  c("tile_id"),
  "defor_tiles"
)


#-----------------------------
# Build canonical tile-year table
# one row per tile-year
#-----------------------------

defor_tile_year <- defor_tabular_raw %>%
  transmute(
    tile_id = as.character(id),
    year = as.integer(lossyear_mode),
    defor_total_ha = as.numeric(total_deforested_ha)
  ) %>%
  filter(!is.na(tile_id), !is.na(year))

assert_no_duplicate_keys(
  defor_tile_year,
  c("tile_id", "year"),
  "defor_tile_year"
)


#-----------------------------
# Canonical cluster site table
# one row per site observation
#-----------------------------
cluster_sites <- cluster_raw %>%
  transmute(
    sample_id = as.character(sample_id),
    AEZ = as.character(AEZ),
    year = as.integer(year),
    cluster_id = as.character(cluster_id),
    ov = as.numeric(ov_score),
    latitude = as.numeric(Latitude),
    longitude = as.numeric(Longitude),
    method = as.character(method),
    dist_to_medoid = as.numeric(dist_to_medoid),
  ) %>%
  filter(
    !is.na(sample_id),
    !is.na(AEZ),
    !is.na(cluster_id),
    !is.na(year),
    !is.na(latitude),
    !is.na(longitude)
  ) %>%
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  ) %>%
  st_transform(analysis_crs)

assert_no_duplicate_keys(
  st_drop_geometry(cluster_sites),
  c("sample_id"),
  "cluster_sites"
)

#-----------------------------
# Canonical cluster-year OV table
# one row per cluster-year
#-----------------------------

cluster_year_ov <- cluster_sites %>%
  st_drop_geometry() %>%
  group_by(AEZ, cluster_id, year) %>%
  summarise(
    median_ov_year = median(ov, na.rm = TRUE),
    mean_ov_year = mean(ov, na.rm = TRUE),
    n_sites_year = n(),
    .groups = "drop"
  ) %>%
  arrange(AEZ, cluster_id, year)

assert_no_duplicate_keys(
  cluster_year_ov,
  c("AEZ", "cluster_id", "year"),
  "cluster_year_ov"
)




input_data <- build_or_load_step(
  "step0_read_input_data",
  {
    defor_spatial <- read_sf(defor_spatial_data) %>%
      st_make_valid() %>%
      mutate(id = trimws(as.character(id)))

    defor_tabular <- read_csv(defor_tabular_data, show_col_types = FALSE) %>%
      mutate(
        id = trimws(as.character(id)),
        lossyear_mode = as.integer(lossyear_mode)
      )

    biodiversity_clusters <- read_csv(cluster_data, show_col_types = FALSE) %>%
      mutate(
        Sample_midpoint = as.Date(Sample_midpoint),
        year = as.integer(year)
      ) %>%
      filter(!is.na(cluster_id)) %>%
      select(
        sample_id,
        Id,
        AEZ,
        Latitude,
        Longitude,
        Sample_midpoint,
        year,
        ov_score,
        cluster_id,
        dist_to_medoid
      )

    list(
      defor_spatial = defor_spatial,
      defor_tabular = defor_tabular,
      biodiversity_clusters = biodiversity_clusters
    )
  }
)

defor_spatial <- input_data$defor_spatial
defor_tabular <- input_data$defor_tabular
biodiversity_clusters <- input_data$biodiversity_clusters
```



## Please try to do something useful dog...
```{r}



```



## Inspection: spatial vs tabular data alignment
```{r}
alignment_outputs <- build_or_load_step(
  "step0_alignment",
  {
    spatial_ids <- unique(defor_spatial$id)
    tabular_ids <- unique(defor_tabular$id)

    ids_only_in_spatial <- setdiff(spatial_ids, tabular_ids)
    ids_only_in_tabular <- setdiff(tabular_ids, spatial_ids)
    shared_ids <- intersect(spatial_ids, tabular_ids)

    spatial_dupes <- defor_spatial %>%
      st_drop_geometry() %>%
      count(id) %>%
      filter(n > 1)

    tabular_dupes <- defor_tabular %>%
      count(id) %>%
      filter(n > 24)

    write_csv(tibble(id = shared_ids), file.path(tmp_dir, "shared_defor_tile_ids.csv"))

    defor_spatial_matched <- defor_spatial %>%
      filter(id %in% shared_ids)

    defor_tabular_matched <- defor_tabular %>%
      filter(id %in% shared_ids)

    list(
      spatial_ids = spatial_ids,
      tabular_ids = tabular_ids,
      ids_only_in_spatial = ids_only_in_spatial,
      ids_only_in_tabular = ids_only_in_tabular,
      shared_ids = shared_ids,
      spatial_dupes = spatial_dupes,
      tabular_dupes = tabular_dupes,
      defor_spatial_matched = defor_spatial_matched,
      defor_tabular_matched = defor_tabular_matched
    )
  }
)

spatial_ids <- alignment_outputs$spatial_ids
tabular_ids <- alignment_outputs$tabular_ids
ids_only_in_spatial <- alignment_outputs$ids_only_in_spatial
ids_only_in_tabular <- alignment_outputs$ids_only_in_tabular
shared_ids <- alignment_outputs$shared_ids
spatial_dupes <- alignment_outputs$spatial_dupes
tabular_dupes <- alignment_outputs$tabular_dupes
defor_spatial_matched <- alignment_outputs$defor_spatial_matched
defor_tabular_matched <- alignment_outputs$defor_tabular_matched

cat("Spatial rows:", nrow(defor_spatial), "\n")
cat("Spatial unique IDs:", length(spatial_ids), "\n")
cat("Tabular rows:", nrow(defor_tabular), "\n")
cat("Tabular unique IDs:", length(tabular_ids), "\n")
cat("Shared IDs:", length(shared_ids), "\n")
cat("IDs only in spatial:", length(ids_only_in_spatial), "\n")
cat("IDs only in tabular:", length(ids_only_in_tabular), "\n")
cat("Spatial IDs with duplicate rows:", nrow(spatial_dupes), "\n")
cat("Tabular IDs with >24 rows:", nrow(tabular_dupes), "\n")
```

## Step 1: prepare deforestation tiles and hectare-info flags
```{r}
step1_outputs <- build_or_load_step(
  "step1_prepare_deforestation_tiles",
  {
    defor_spatial_unique <- defor_spatial_matched %>%
      group_by(id) %>%
      summarise(geometry = st_union(geometry), .groups = "drop")

    tile_year_coverage <- defor_tabular_matched %>%
      count(id, name = "n_years")

    complete_tile_ids <- tile_year_coverage %>%
      filter(n_years == 24) %>%
      pull(id)

    defor_tiles_all_sf <- defor_spatial_unique %>%
      mutate(
        has_ha_info = id %in% complete_tile_ids,
        tile_fill_color = if_else(has_ha_info, "darkgreen", "tomato")
      )

    defor_tiles_complete_sf <- defor_tiles_all_sf %>%
      filter(has_ha_info)

    list(
      defor_spatial_unique = defor_spatial_unique,
      tile_year_coverage = tile_year_coverage,
      complete_tile_ids = complete_tile_ids,
      defor_tiles_all_sf = defor_tiles_all_sf,
      defor_tiles_complete_sf = defor_tiles_complete_sf
    )
  }
)

defor_spatial_unique <- step1_outputs$defor_spatial_unique
tile_year_coverage <- step1_outputs$tile_year_coverage
complete_tile_ids <- step1_outputs$complete_tile_ids
defor_tiles_all_sf <- step1_outputs$defor_tiles_all_sf
defor_tiles_complete_sf <- step1_outputs$defor_tiles_complete_sf

cat("All matched tiles:", n_distinct(defor_tiles_all_sf$id), "\n")
cat("Tiles with hectare info:", n_distinct(defor_tiles_complete_sf$id), "\n")
```

## Step 2: build cluster-level OV summaries
```{r}
step2_outputs <- build_or_load_step(
  "step2_cluster_ov_summaries",
  {
    cluster_year_ov <- biodiversity_clusters %>%
      group_by(AEZ, cluster_id, year) %>%
      summarise(
        median_ov_year = median(ov_score, na.rm = TRUE),
        n_sites_year = n(),
        .groups = "drop"
      ) %>%
      arrange(AEZ, cluster_id, year)

    cluster_ov_trend <- cluster_year_ov %>%
      group_by(AEZ, cluster_id) %>%
      summarise(
        first_year = first(year),
        last_year = last(year),
        ov_first = first(median_ov_year),
        ov_last = last(median_ov_year),
        ov_change = ov_last - ov_first,
        year_span = last_year - first_year,
        n_ov_years = n_distinct(year),
        .groups = "drop"
      ) %>%
      mutate(
        ov_trend = case_when(
          abs(ov_change) < 0.1 ~ "OV_no_change",
          ov_change < 0 ~ "OV_declines",
          ov_change > 0 ~ "OV_increases",
          TRUE ~ "OV_no_change"
        ),
        medoid_color = case_when(
          ov_trend == "OV_declines" ~ "green3",
          ov_trend == "OV_increases" ~ "red2",
          TRUE ~ "grey40"
        )
      )

    list(
      cluster_year_ov = cluster_year_ov,
      cluster_ov_trend = cluster_ov_trend
    )
  }
)

cluster_year_ov <- step2_outputs$cluster_year_ov
cluster_ov_trend <- step2_outputs$cluster_ov_trend

cluster_ov_trend
```

## Step 3: build cluster medoid points and site points
```{r}
step3_outputs <- build_or_load_step(
  "step3_spatial_cluster_objects",
  {
    cluster_medoids <- biodiversity_clusters %>%
      group_by(AEZ, cluster_id) %>%
      slice_min(order_by = dist_to_medoid, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      left_join(cluster_ov_trend, by = c("AEZ", "cluster_id"))

    cluster_medoids_sf <- cluster_medoids %>%
      filter(!is.na(Longitude), !is.na(Latitude)) %>%
      st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)

    sites_sf_all <- biodiversity_clusters %>%
      filter(!is.na(Longitude), !is.na(Latitude)) %>%
      st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE) %>%
      st_transform(analysis_crs)

    defor_tiles_all_proj <- defor_tiles_all_sf %>%
      st_transform(analysis_crs)

    defor_tiles_complete_proj <- defor_tiles_complete_sf %>%
      st_transform(analysis_crs)

    list(
      cluster_medoids = cluster_medoids,
      cluster_medoids_sf = cluster_medoids_sf,
      sites_sf_all = sites_sf_all,
      defor_tiles_all_proj = defor_tiles_all_proj,
      defor_tiles_complete_proj = defor_tiles_complete_proj
    )
  }
)

cluster_medoids <- step3_outputs$cluster_medoids
cluster_medoids_sf <- step3_outputs$cluster_medoids_sf
sites_sf_all <- step3_outputs$sites_sf_all
defor_tiles_all_proj <- step3_outputs$defor_tiles_all_proj
defor_tiles_complete_proj <- step3_outputs$defor_tiles_complete_proj
```

## Step 4: build cluster footprints and tag them to deforestation tiles
```{r}
step4_outputs <- build_or_load_step(
  "step4_cluster_tile_tagging",
  {
    cluster_analysis_outputs <- purrr::map(
      buffer_km_vals,
      function(buffer_km) {
        buffer_m <- buffer_km * 1000

        cluster_footprints <- sites_sf_all %>%
          mutate(buf_geom = st_buffer(geometry, dist = buffer_m)) %>%
          st_set_geometry("buf_geom") %>%
          group_by(AEZ, cluster_id) %>%
          summarise(
            n_sites = n(),
            geometry = st_union(geometry),
            .groups = "drop"
          )

        intersects_all <- st_intersects(cluster_footprints, defor_tiles_all_proj)
        intersects_complete <- st_intersects(cluster_footprints, defor_tiles_complete_proj)

        cluster_footprints_tagged <- cluster_footprints %>%
          mutate(
            buffer_km = buffer_km,
            tagged_any_tile = lengths(intersects_all) > 0,
            tagged_ha_tile = lengths(intersects_complete) > 0
          )

        cluster_tile_matches <- purrr::map2_dfr(
          seq_len(nrow(cluster_footprints_tagged)),
          intersects_all,
          function(row_idx, tile_idx) {
            footprint_row <- cluster_footprints_tagged[row_idx, ] %>%
              st_drop_geometry()

            if (length(tile_idx) == 0) {
              return(
                tibble(
                  AEZ = footprint_row$AEZ,
                  cluster_id = footprint_row$cluster_id,
                  n_sites = footprint_row$n_sites,
                  buffer_km = footprint_row$buffer_km,
                  tagged_any_tile = footprint_row$tagged_any_tile,
                  tagged_ha_tile = footprint_row$tagged_ha_tile,
                  tile_id = NA_character_,
                  has_ha_info = NA
                )
              )
            }

            tibble(
              AEZ = footprint_row$AEZ,
              cluster_id = footprint_row$cluster_id,
              n_sites = footprint_row$n_sites,
              buffer_km = footprint_row$buffer_km,
              tagged_any_tile = footprint_row$tagged_any_tile,
              tagged_ha_tile = footprint_row$tagged_ha_tile,
              tile_id = defor_tiles_all_proj$id[tile_idx],
              has_ha_info = defor_tiles_all_proj$has_ha_info[tile_idx]
            )
          }
        )

        cluster_year_defor <- cluster_tile_matches %>%
          filter(!is.na(tile_id), has_ha_info) %>%
          select(AEZ, cluster_id, buffer_km, tile_id) %>%
          distinct() %>%
          left_join(
            defor_tabular_matched %>%
              rename(tile_id = id, year = lossyear_mode),
            by = "tile_id",
            relationship = "many-to-many"
          ) %>%
          group_by(AEZ, cluster_id, buffer_km, year) %>%
          summarise(
            defor_total_ha_year = sum(total_deforested_ha, na.rm = TRUE),
            n_tiles_with_ha = n_distinct(tile_id),
            .groups = "drop"
          ) %>%
          arrange(AEZ, cluster_id, buffer_km, year)

        list(
          footprints = cluster_footprints_tagged,
          tile_matches = cluster_tile_matches,
          year_defor = cluster_year_defor
        )
      }
    )

    names(cluster_analysis_outputs) <- paste0("buffer_", buffer_km_vals, "km")

    list(
      cluster_analysis_outputs = cluster_analysis_outputs,
      cluster_footprints_all = purrr::map_dfr(cluster_analysis_outputs, "footprints"),
      cluster_tile_matches_all = purrr::map_dfr(cluster_analysis_outputs, "tile_matches"),
      cluster_year_defor_all = purrr::map_dfr(cluster_analysis_outputs, "year_defor")
    )
  }
)

cluster_analysis_outputs <- step4_outputs$cluster_analysis_outputs
cluster_footprints_all <- step4_outputs$cluster_footprints_all
cluster_tile_matches_all <- step4_outputs$cluster_tile_matches_all
cluster_year_defor_all <- step4_outputs$cluster_year_defor_all
```

## Step 5: create per-transition reporting table
```{r}
## Step 5: create per-transition reporting table
step5_outputs <- build_or_load_step(
  "step5_cluster_reports",
  {
    cluster_year_pairs <- cluster_year_ov %>%
      group_by(AEZ, cluster_id) %>%
      arrange(year, .by_group = TRUE) %>%
      mutate(
        year_t2 = lead(year),
        ov_t2 = lead(median_ov_year)
      ) %>%
      ungroup() %>%
      filter(!is.na(year_t2)) %>%
      transmute(
        AEZ,
        cluster_id,
        year_t1 = year,
        year_t2,
        year_gap = year_t2 - year_t1,
        ov_t1 = median_ov_year,
        ov_t2,
        delta_ov = ov_t2 - ov_t1
      )

    cluster_transition_keys <- cluster_footprints_all %>%
      st_drop_geometry() %>%
      select(
        AEZ,
        cluster_id,
        buffer_km,
        n_sites,
        tagged_any_tile,
        tagged_ha_tile
      ) %>%
      distinct() %>%
      left_join(
        cluster_tile_matches_all %>%
          group_by(AEZ, cluster_id, buffer_km) %>%
          summarise(
            n_matched_tiles = n_distinct(tile_id[!is.na(tile_id)]),
            n_matched_tiles_with_ha = n_distinct(tile_id[has_ha_info %in% TRUE]),
            .groups = "drop"
          ),
        by = c("AEZ", "cluster_id", "buffer_km")
      )

    defor_t1_lookup <- cluster_year_defor_all %>%
      transmute(
        AEZ,
        cluster_id,
        buffer_km,
        year_t1 = year,
        defor_ha_t1 = defor_total_ha_year
      )

    defor_t2_lookup <- cluster_year_defor_all %>%
      transmute(
        AEZ,
        cluster_id,
        buffer_km,
        year_t2 = year,
        defor_ha_t2 = defor_total_ha_year
      )

    stopifnot(
      !anyDuplicated(cluster_transition_keys[c("AEZ", "cluster_id", "buffer_km")]),
      !anyDuplicated(cluster_year_pairs[c("AEZ", "cluster_id", "year_t1", "year_t2")]),
      !anyDuplicated(defor_t1_lookup[c("AEZ", "cluster_id", "buffer_km", "year_t1")]),
      !anyDuplicated(defor_t2_lookup[c("AEZ", "cluster_id", "buffer_km", "year_t2")])
    )

    cluster_pair_report <- cluster_transition_keys %>%
      inner_join(
        cluster_year_pairs,
        by = c("AEZ", "cluster_id"),
        relationship = "many-to-many"
      ) %>%
      left_join(
        defor_t1_lookup,
        by = c("AEZ", "cluster_id", "buffer_km", "year_t1")
      ) %>%
      left_join(
        defor_t2_lookup,
        by = c("AEZ", "cluster_id", "buffer_km", "year_t2")
      ) %>%
      mutate(
        delta_defor_ha = defor_ha_t2 - defor_ha_t1,
        delta_ov_annualized = delta_ov / year_gap,
        delta_defor_annualized = delta_defor_ha / year_gap,
        favorable_case = !is.na(delta_ov) & !is.na(delta_defor_ha) &
          delta_ov > 0 & delta_defor_ha < 0,
        unfavorable_case = !is.na(delta_ov) & !is.na(delta_defor_ha) &
          delta_ov < 0 & delta_defor_ha > 0,
        opposing_change = favorable_case | unfavorable_case
      ) %>%
      select(
        AEZ,
        cluster_id,
        buffer_km,
        n_sites,
        n_matched_tiles,
        n_matched_tiles_with_ha,
        year_t1,
        year_t2,
        year_gap,
        ov_t1,
        ov_t2,
        delta_ov,
        defor_ha_t1,
        defor_ha_t2,
        delta_defor_ha,
        favorable_case
      ) %>%
      arrange(AEZ, cluster_id, buffer_km, year_t1, year_t2)

    list(
      cluster_year_pairs = cluster_year_pairs,
      cluster_transition_keys = cluster_transition_keys,
      cluster_pair_report = cluster_pair_report
    )
  }
)

cluster_year_pairs <- step5_outputs$cluster_year_pairs
cluster_report_base <- step5_outputs$cluster_report_base
cluster_pair_report <- step5_outputs$cluster_pair_report
cluster_case_summary <- step5_outputs$cluster_case_summary
```

## Step 6: summarize tagging counts by AEZ and buffer size
```{r}
step6_outputs <- build_or_load_step(
  "step6_tag_counts",
  {
    cluster_tag_counts <- cluster_footprints_all %>%
      st_drop_geometry() %>%
      count(AEZ, buffer_km, tagged_any_tile, name = "n_clusters") %>%
      mutate(
        tag_status = if_else(tagged_any_tile, "tagged", "not_tagged")
      ) %>%
      select(AEZ, buffer_km, tag_status, n_clusters)

    list(cluster_tag_counts = cluster_tag_counts)
  }
)

cluster_tag_counts <- step6_outputs$cluster_tag_counts
```

## Step 7: save analysis outputs
```{r}
write_csv(
  cluster_tag_counts,
  file.path(out_dir, "clusters_tag_counts.csv")
)

write_csv(
  cluster_pair_report,
  file.path(out_dir, "cluster_pair_report.csv")
)

write_csv(
  cluster_case_summary,
  file.path(out_dir, "cluster_case_summary.csv")
)

if (file.exists(gpkg_path)) file.remove(gpkg_path)

st_write(
  cluster_medoids_sf,
  gpkg_path,
  layer = "cluster_medoids",
  delete_layer = TRUE,
  quiet = TRUE
)

st_write(
  defor_tiles_all_sf,
  gpkg_path,
  layer = "defor_tiles",
  delete_layer = TRUE,
  quiet = TRUE
)

purrr::walk(
  buffer_km_vals,
  function(buffer_value) {
    st_write(
      cluster_footprints_all %>%
        filter(buffer_km == buffer_value),
      gpkg_path,
      layer = paste0("cluster_fp_", buffer_value, "km"),
      delete_layer = TRUE,
      quiet = TRUE
    )
  }
)
```

### Visualization and comparison

## Plot medoids and deforestation tile coverage
```{r}
world_map <- map_data("world")

ggplot() +
  geom_polygon(
    data = world_map,
    aes(x = long, y = lat, group = group),
    fill = "grey95",
    color = "grey80",
    linewidth = 0.2
  ) +
  geom_sf(
    data = defor_tiles_all_sf,
    aes(fill = has_ha_info),
    color = NA,
    alpha = 0.30,
    inherit.aes = FALSE
  ) +
  geom_sf(
    data = cluster_medoids_sf,
    aes(color = ov_trend),
    size = 1.6,
    alpha = 0.9,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(
    values = c(`TRUE` = "darkgreen", `FALSE` = "tomato"),
    name = "Tile has\nha info"
  ) +
  scale_color_manual(
    values = c(
      OV_declines = "green3",
      OV_increases = "red2",
      OV_no_change = "grey40"
    ),
    name = "OV trend"
  ) +
  coord_sf(expand = FALSE) +
  theme_minimal() +
  labs(
    title = "Cluster medoids and deforestation tile coverage",
    subtitle = "Medoids colored by first-to-last OV change: decline, increase, or near-no-change (< 0.1)"
  )
```

## Comparison tables
```{r}
cluster_tag_counts
cluster_case_summary
```
