---
title: "R Notebook"
author: "Stanley Tan (original), Eli Duggan (edits)"
output: html_document
---

## Libraries and helper functions
```{r}
library(ape)
library(picante)
library(tidyverse)
library(ggplot2)
library(sf)

ensure_dir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
  invisible(path)
}

minmax01 <- function(x) {
  r <- range(x, na.rm = TRUE)
  (x - r[1]) / (r[2] - r[1])
}
```

## Configuration
```{r}
# Directories
tmp_dir <- file.path("..", "tmp")
out_dir <- file.path("..", "output")
in_dir <- file.path("..", "input")
build_out <- file.path("..", "..", "build", "output")
dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)

# Copy output files from "build"
files_to_copy <- c("predicts_filtered_2000_2024.csv", "pd_result.csv")

walk(files_to_copy, \(f) {
  file.copy(
    from = file.path(build_out, f),
    to   = file.path(in_dir, f),
    overwrite = TRUE
  )
})

# File paths
predicts_file <- file.path(in_dir, "predicts_filtered_2000_2024.csv")
pd_file       <- file.path(in_dir, "pd_result.csv")
```

## Read input data
```{r}
predicts <- read_csv(predicts_file, show_col_types = FALSE) %>%
  mutate(
    Sample_midpoint = as.Date(Sample_midpoint),
    sample_id = paste(SSBS, Sample_midpoint, sep = "__")
  )
pd       <- read_csv(pd_file, show_col_types = FALSE)
```

# Calculating subcomponents: Shannon index, MSA, HQI, HANPP
```{r}
# reusable abundance by taxon block
taxon_ab <- predicts %>%
  filter(Effort_corrected_measurement > 0) %>%
  mutate(taxon = paste(Family, Genus, sep = "_")) %>%
  group_by(sample_id, taxon) %>%
  summarise(ab = sum(Effort_corrected_measurement), .groups = "drop") %>%
  mutate(ab = round(ab, 0))

## Shannon index (species count + evenness) at sample_id-level
shann_index <- taxon_ab %>%
  group_by(sample_id) %>%
  mutate(p = ab / sum(ab)) %>%
  summarise(shannon = -sum(p * log(p), na.rm = TRUE), .groups = "drop")
print("Shannon index calculated.")

## Mean Species Abundance (MSA) at sample_id-level
msa <- taxon_ab %>%
  group_by(sample_id) %>%
  summarise(
    site_total_ab  = sum(ab),
    site_total_spp = n_distinct(taxon),
    msa = round(site_total_ab / site_total_spp, 0),
    .groups = "drop"
  )
print("MSA calculated.")

## Habitat Quality Index at sample_id-level
hqi_score <- predicts %>%
  distinct(sample_id, Biome, Predominant_land_use, Use_intensity) %>%
  filter(Predominant_land_use != "Cannot decide",
         Use_intensity != "Cannot decide") %>%
  mutate(
    lu_score = if_else(
      Predominant_land_use %in% c(
        "Mature secondary vegetation",
        "Secondary vegetation (indeterminate age)",
        "Intermediate secondary vegetation",
        "Primary vegetation",
        "Young secondary vegetation"
      ),
      3, 1
    ),
    ui_score = case_when(
      Use_intensity == "Minimal use" ~ -0.1,
      Use_intensity == "Light use"   ~ -0.3,
      Use_intensity == "Intense use" ~ -0.5,
      TRUE ~ NA_real_
    ),
    hq_score = lu_score + ui_score
  ) %>%
  select(sample_id, hq_score)
print("HQI calculated.")

## HANPP
hanpp_score <- predicts %>%
  distinct(sample_id, Biome, Predominant_land_use, Use_intensity) %>%
  mutate(
    hanpp = case_when(
      Predominant_land_use == "Urban" ~ "high",
      Predominant_land_use %in% c(
        "Primary vegetation",
        "Mature secondary vegetation",
        "Intermediate secondary vegetation",
        "Young secondary vegetation",
        "Secondary vegetation (indeterminate age)"
      ) ~ "low",
      Predominant_land_use %in% c("Cropland", "Plantation forest", "Pasture") &
        Use_intensity == "Intense use" ~ "high",
      Predominant_land_use %in% c("Cropland", "Plantation forest", "Pasture") &
        Use_intensity %in% c("Minimal use", "Light use") ~ "low",
      Predominant_land_use == "Cannot decide" | Use_intensity == "Cannot decide" ~ NA_character_,
      TRUE ~ NA_character_
    ),
    hanpphigh = as.integer(hanpp == "high"),
    hanpplow  = as.integer(hanpp == "low")
  ) %>%
  select(sample_id, hanpp, hanpphigh, hanpplow)
print("HANPP calculated.")


```

## calculating PDF (sample_id by Class-level)
```{r}
# Define land use sets
baseline_primary <- "Primary vegetation"
baseline_secondary <- c(
  "Young secondary vegetation",
  "Intermediate secondary vegetation",
  "Mature secondary vegetation",
  "Secondary vegetation (indeterminate age)"
)
impact_uses <- c("Cropland", "Pasture", "Plantation forest", "Urban")

# Resolve a single biome per sample_id (most frequent biome label)
biome_lookup <- predicts %>%
  filter(!is.na(sample_id), !is.na(Biome)) %>%
  count(sample_id, Biome, name = "n") %>%
  slice_max(order_by = n, n = 1, by = sample_id, with_ties = FALSE) %>%
  select(sample_id, Biome) %>%
  rename(Biome_resolved = Biome)

# 1) Aggregate abundance to sample_id × Class × Biome_resolved × land-use
pdf_base <- predicts %>%
  select(sample_id, Class, Predominant_land_use, Effort_corrected_measurement) %>%
  left_join(biome_lookup, by = "sample_id") %>%
  filter(
    !is.na(sample_id),
    !is.na(Class),
    !is.na(Biome_resolved),
    Predominant_land_use != "Cannot decide",
    Effort_corrected_measurement > 0
  ) %>%
  group_by(sample_id, Class, Biome_resolved, Predominant_land_use) %>%
  summarise(total_ab = sum(Effort_corrected_measurement), .groups = "drop") %>%
  mutate(total_ab = round(total_ab, 2))

# 2) Compute baseline abundance for each sample_id × Class × Biome_resolved
baseline_tbl <- pdf_base %>%
  mutate(
    baseline_type = case_when(
      Predominant_land_use == baseline_primary ~ "primary_b",
      Predominant_land_use %in% baseline_secondary ~ "secondary_b",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(baseline_type)) %>%
  group_by(sample_id, Class, Biome_resolved, baseline_type) %>%
  summarise(baseline = mean(total_ab, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = baseline_type, values_from = baseline)

# 3) Get impacted land-use totals per sample_id × Class × Biome_resolved
impact_tbl <- pdf_base %>%
  filter(Predominant_land_use %in% impact_uses) %>%
  select(sample_id, Class, Biome_resolved, Predominant_land_use, total_ab) %>%
  pivot_wider(
    names_from = Predominant_land_use,
    values_from = total_ab,
    values_fill = 0
  )

# 4) Join and compute percent changes vs primary and secondary baselines
#    (epsilon avoids division-by-zero blowups)
eps <- 1e-7

pdf_component <- baseline_tbl %>%
  left_join(impact_tbl, by = c("sample_id", "Class", "Biome_resolved")) %>%
  # keep rows that have at least one baseline
  filter(!(is.na(primary_b) & is.na(secondary_b))) %>%
  mutate(
    cropland_pri   = 100 * (Cropland - primary_b) / (primary_b + eps),
    pasture_pri    = 100 * (Pasture - primary_b) / (primary_b + eps),
    plantation_pri = 100 * (`Plantation forest` - primary_b) / (primary_b + eps),
    urban_pri      = 100 * (Urban - primary_b) / (primary_b + eps),

    cropland_sec   = 100 * (Cropland - secondary_b) / (secondary_b + eps),
    pasture_sec    = 100 * (Pasture - secondary_b) / (secondary_b + eps),
    plantation_sec = 100 * (`Plantation forest` - secondary_b) / (secondary_b + eps),
    urban_sec      = 100 * (Urban - secondary_b) / (secondary_b + eps)
  ) %>%
  mutate(across(
    c(cropland_pri, pasture_pri, plantation_pri, urban_pri,
      cropland_sec, pasture_sec, plantation_sec, urban_sec),
    ~ round(., 0)
  )) %>%
  mutate(
    mean_avg = rowMeans(
      select(., cropland_pri, pasture_pri, plantation_pri, urban_pri,
                cropland_sec, pasture_sec, plantation_sec, urban_sec),
      na.rm = TRUE
    ),
    mean_avg = ifelse(is.nan(mean_avg), 0, mean_avg)
  ) %>%
  rename(Biome = Biome_resolved) %>%
  select(
    sample_id, Class, Biome,
    primary_b, secondary_b,
    Cropland, Pasture, `Plantation forest`, Urban,
    cropland_pri, pasture_pri, plantation_pri, urban_pri,
    cropland_sec, pasture_sec, plantation_sec, urban_sec,
    mean_avg
  )

# 5) Integrity check: one row per sample_id × Class
stopifnot(
  nrow(pdf_component) ==
    n_distinct(paste(pdf_component$sample_id, pdf_component$Class))
)

print("PDF calculated.")
```

## Prepare to assemble table of calculated OV components (PD, HANPP, MSA, Shannon Index, PDF, HQI)
```{r}
# Lookups for date and coordinates
site_date_lookup <- predicts %>%
  distinct(sample_id, SSBS, Sample_midpoint)

coords_lookup <- predicts %>%
  filter(!is.na(SSBS), !is.na(Latitude), !is.na(Longitude)) %>%
  distinct(SSBS, Latitude, Longitude)

# quality control coordinates - check each SSBS has one coordinate pair
stopifnot(coords_lookup %>% count(SSBS) %>% summarise(all(n == 1)) %>% pull())

# because PDF is calculated at class-level, we need to bring it to site-level to align with the rest of the OV components.
pdf_sample <- pdf_component %>%
  group_by(sample_id) %>%
  summarise(mean_pct_change = mean(mean_avg, na.rm = TRUE), .groups = "drop")
```

# Assemble OV component table (site-level)
```{r}
combined_component_scores <- pd %>%   # <-- use pd (what you actually read)
  select(sample_id, phylo_div, SR) %>%
  left_join(pdf_sample,  by = "sample_id") %>%
  left_join(shann_index, by = "sample_id") %>%
  left_join(msa,         by = "sample_id") %>%
  left_join(hqi_score,   by = "sample_id") %>%
  left_join(hanpp_score, by = "sample_id") %>%
  left_join(site_date_lookup, by = "sample_id") %>%
  left_join(coords_lookup,    by = "SSBS") %>%
  filter(!is.na(hanpp), !is.na(shannon), !is.nan(shannon))

# QC: uniqueness (one row per sample_id)
stopifnot(nrow(combined_component_scores) == n_distinct(combined_component_scores$sample_id))

combined_out <- file.path(tmp_dir, "combined_component_scores_sitelevel.csv")
write_csv(combined_component_scores, combined_out)
```

## Calculating OV score with scaled indicate
```{r}
# read in if needed
# combined_component_scores <- read.csv("combined_component_scores_sitelevel.csv")

model_df <- combined_component_scores %>%
  mutate(
    log_msa = log(msa + 1),
    shannon_scaled = minmax01(shannon),
    phylo_scaled   = minmax01(phylo_div),
    log_msa_scaled = minmax01(log_msa),
    ov_score =
      hq_score +
      shannon_scaled +
      phylo_scaled +
      log_msa_scaled +
      0.2 * hanpphigh +
      0.8 * hanpplow
  ) %>%
  select(
    sample_id, Latitude, Longitude, Sample_midpoint,
    hanpp, msa, hq_score, shannon, phylo_div,
    hanpphigh, hanpplow, log_msa,
    shannon_scaled, phylo_scaled, log_msa_scaled, ov_score
  )

ov_out <- file.path(tmp_dir, "ov_scores_sitelevel.csv")
write_csv(model_df, ov_out)
```

## Assign AEZ
```{r}
sf_use_s2(FALSE)  # avoid polygon validity checks


aez_path <- file.path("..","..","..","spatial_data", "aez", "AEZ_shp_file.shp")
if (!file.exists(aez_path)) {
  stop("AEZ shapefile not found at: ", normalizePath(aez_path, mustWork = FALSE))
}

aez <- read_sf(aez_path) %>%
  st_make_valid() %>%
  select(Id, AEZ)

sites_sf <- model_df %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE) %>%
  st_transform(st_crs(aez))

# strict tagging
tagged <- st_join(sites_sf, aez, join = st_within, left = TRUE) %>%
  st_drop_geometry()

# nearest fallback for unassigned
missing_idx <- which(is.na(tagged$AEZ))
if (length(missing_idx) > 0) {
  nearest_idx <- st_nearest_feature(sites_sf[missing_idx, ], aez)
  tagged$AEZ[missing_idx] <- aez$AEZ[nearest_idx]
}

tagged <- tagged %>%
  mutate(AEZ_assigned_by_nearest = row_number() %in% missing_idx) %>%
  relocate(AEZ, .after = sample_id) %>%
  relocate(AEZ_assigned_by_nearest, .after = AEZ)

aez_out <- file.path(out_dir, "ov_AEZ_tag.csv")
write_csv(tagged, aez_out)
```

## Optional diagnostic - Compare within and intersects for duplicates/missing assignments
```{r}
# I compared sf_within and sf_intersects to see which produces duplicates or untagged sites --> sf_within has no duplicates, so using it

tag_within <- st_join(sites_sf, aez, join = st_within, left = TRUE) %>%
  st_drop_geometry() %>%
  group_by(sample_id) %>%
  slice(1) %>%
  ungroup() %>%
  select(sample_id, AEZ_within = AEZ)

tag_intersects <- st_join(sites_sf, aez, join = st_intersects, left = TRUE) %>%
  st_drop_geometry() %>%
  group_by(sample_id) %>%
  slice(1) %>%
  ungroup() %>%
  select(sample_id, AEZ_intersects = AEZ)

compare <- full_join(tag_within, tag_intersects, by = "sample_id")

compare_summary <- compare %>%
  summarise(
    n_total = n(),
    n_equal = sum(AEZ_within == AEZ_intersects, na.rm = TRUE),
    n_diff  = sum(AEZ_within != AEZ_intersects, na.rm = TRUE),
    n_within_only     = sum(!is.na(AEZ_within) & is.na(AEZ_intersects)),
    n_intersects_only = sum(is.na(AEZ_within) & !is.na(AEZ_intersects)),
    n_both_na         = sum(is.na(AEZ_within) & is.na(AEZ_intersects))
  )

print(compare_summary)

compare_diffs <- compare %>%
  filter(!is.na(AEZ_within), !is.na(AEZ_intersects), AEZ_within != AEZ_intersects)

# If you want to inspect:
# print(compare_diffs)

# ----------------------------
# Quick counts / map (optional)
# ----------------------------
# ggplot(aez) + geom_sf()
# table(tagged$AEZ)
```