---
title: "biodiversity_site_clustering"
author: "Eli Duggan"
date: "2026-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
library(sf)
library(dplyr)
library(cluster)
library(tidyverse)
library(purrr)
library(furrr)
library(progressr)
library(future)
library(ggplot2)
library(here)
library(stringr)
library(tidyr)
library(readr)
library(sf)
```

## Libraries and helper functions
```{r}
ensure_dir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
  invisible(path)
}
```

## Configuration
```{r}
tmp_dir <- file.path("..", "tmp")
out_dir <- file.path("..", "output")
in_dir <- file.path("..", "input")
predicts_out_dir <- file.path("..", "..", "predicts_ov_table", "analysis", "output")
aez_dir <- file.path("..", "..", "spatial_data", "aez")
checkpoint_root <- file.path(tmp_dir, "checkpoints")
summary_dir <- file.path(out_dir, "summary")

walk(c(tmp_dir, out_dir, in_dir, checkpoint_root, summary_dir), ensure_dir)

files_to_copy <- c("ov_AEZ_tag.csv")

walk(files_to_copy, \(f) {
  file.copy(
    from = file.path(predicts_out_dir, f),
    to = file.path(in_dir, f),
    overwrite = TRUE
  )
})

tagged_sites_file <- file.path(in_dir, "ov_AEZ_tag.csv")
aez_path <- file.path(aez_dir, "AEZ_shp_file.shp")
```

## Read input data
```{r}
# Read site observations and standardize coordinates and year
model_df_tagged <- read_csv(tagged_sites_file) %>%
  mutate(
    lat_r = round(Latitude, 4),
    lon_r = round(Longitude, 4),
    year = as.integer(substr(Sample_midpoint, 1, 4))
  )

# bar chart - number of years covered
#   - see pretty decent year coverage, actually
ggplot(model_df_tagged, aes(x = year)) +
  geom_bar() +
  labs(x = "Year", y = "Number of site-year locations", title = "Model_df_tagged")

# find unique site locations to reduce clustering burden
#   - keeps one record per AEZ, per rounded location
unique_sites <- model_df_tagged %>%
  distinct(AEZ, lat_r, lon_r, .keep_all = TRUE) %>% 
  arrange(AEZ)

# convert to sf points in geographic coordinates (WGS84, EPSG 4326); project onto meter-based CRS (6933 for WEC); pull projected coordinates into plain columns
unique_sf <- unique_sites %>%
  st_as_sf(coords = c("lon_r", "lat_r"), crs = 4326, remove = FALSE, na.fail = FALSE)

sites_m <- st_transform(unique_sf, 6933)

xy <- st_coordinates(sites_m)

sites_tbl <- sites_m %>%
  st_drop_geometry() %>%
  mutate(
    AEZ = as.character(AEZ),
    x = xy[,1],
    y = xy[,2])

# check: dropping geometry doesn't impact row count
stopifnot(nrow(sites_tbl) == nrow(unique_sites))

# compute the year coverage per site, join into sites_tbl, pare down to needed columns 
site_years <- model_df_tagged %>% 
  group_by(AEZ, lat_r, lon_r) %>% 
  summarize(
    n_years = n_distinct(year),
    years_single = if_else(n_years == 1L, first(year), NA_integer_),
    .groups = "drop"
  )

sites_tbl <- sites_tbl %>%
  left_join(site_years, by = c("AEZ","lat_r","lon_r")) %>% 
  select(AEZ, lat_r, lon_r, x, y, n_years, years_single)

# visualize year coverage
ggplot(sites_tbl, aes(x = factor(n_years))) +
  geom_bar() +
  labs(
    x = "Years of data", 
    y = "Number of locations", 
    title = "Years of data at each location"
  )
table(sites_tbl$n_years)

```

## Clustering engines
```{r}
# run the greedy cover
engine_greedy_cover <- function(df, radius_m = 12500) {
  stopifnot(all(c("x","y") %in% names(df)))
  n <- nrow(df)
  x <- df$x; y <- df$y

  cluster <- integer(n)
  dist_to_medoid <- rep(NA_real_, n)

  uncovered <- rep(TRUE, n)
  k <- 0L

  for (i in seq_len(n)) {
    if (!uncovered[i]) next
    k <- k + 1L

    # ball around i
    d <- dist2(x, y, x[i], y[i])
    members <- which(uncovered & d <= radius_m)

    # pick medoid inside members, then re-ball around medoid
    med <- choose_medoid(x, y, members)
    d2 <- dist2(x, y, x[med], y[med])
    members2 <- which(uncovered & d2 <= radius_m)

    cluster[members2] <- k
    dist_to_medoid[members2] <- d2[members2]
    uncovered[members2] <- FALSE
  }

  list(
    cluster = cluster,
    dist_to_medoid = dist_to_medoid,
    k = if (k > 0) k else 0L,
    max_cluster_radius = if (k > 0) max(dist_to_medoid, na.rm = TRUE) else 0,
    method = "greedy_cover",
    params = list(radius_m = radius_m)
  )
}

# Run PAM with max radius constraint
engine_pam_radius <- function(df, radius_m = 12500, k_start = 2L, k_max = 200L) {
  stopifnot(all(c("x","y") %in% names(df)))
  n <- nrow(df)
  x <- df$x; y <- df$y

  if (n <= 1) {
    return(list(
      cluster = if (n == 1) 1L else integer(0),
      dist_to_medoid = if (n == 1) 0 else numeric(0),
      k = if (n == 1) 1L else 0L,
      max_cluster_radius = 0,
      method = "pam_radius",
      params = list(radius_m = radius_m, k_start = k_start, k_max = k_max)
    ))
  }

  # full distance matrix (expensive)
  D <- as.matrix(stats::dist(cbind(x, y), method = "euclidean"))

  best <- NULL
  for (k in seq(from = k_start, to = min(k_max, n))) {
    pam_res <- cluster::pam(D, k = k, diss = TRUE)

    cl <- as.integer(pam_res$clustering)
    medoid_idx <- as.integer(pam_res$id.med)

    dist_to_medoid <- rep(NA_real_, n)
    max_rad <- 0
    for (j in seq_len(k)) {
      idx <- which(cl == j)
      med <- medoid_idx[j]
      d <- D[idx, med]
      dist_to_medoid[idx] <- d
      max_rad <- max(max_rad, max(d))
    }

    best <- list(
      cluster = cl,
      dist_to_medoid = dist_to_medoid,
      k = k,
      max_cluster_radius = max_rad,
      method = "pam_radius",
      params = list(radius_m = radius_m, k_start = k_start, k_max = k_max)
    )

    if (max_rad <= radius_m) break
  }

  best
}

# run CLARA, increasing k until radius constraint is passed
engine_clara <- function(df,
                               radius_m = 12500,
                               k_start = 2L,
                               k_max = 500L,
                               samples = 5L,
                               sampsize = NULL,
                               seed = 1L) {
  stopifnot(all(c("x","y") %in% names(df)))
  n <- nrow(df)
  X <- as.matrix(df[, c("x","y")])

  if (n <= 1) {
    return(list(
      cluster = if (n == 1) 1L else integer(0),
      dist_to_medoid = if (n == 1) 0 else numeric(0),
      k = if (n == 1) 1L else 0L,
      max_cluster_radius = 0,
      method = "clara_radius",
      params = list(radius_m = radius_m, k_start = k_start, k_max = k_max,
                    samples = samples, sampsize = sampsize, seed = seed)
    ))
  }

  if (is.null(sampsize)) {
    # cluster::clara default is min(n, 40 + 2k) historically; we choose something stable
    sampsize <- min(n, max(40L, 10L * k_start))
  }

  best <- NULL

  for (k in seq(from = k_start, to = min(k_max, n))) {
    set.seed(seed)  # deterministic given seed
    clara_res <- cluster::clara(X, k = k, samples = samples, sampsize = sampsize)

    # clara_res$medoids is a k x 2 matrix of medoid coordinates
    med_xy <- as.matrix(clara_res$medoids)

    # Assign all points to nearest medoid + distance (pure geometric)
    assigned <- assign_to_medoids(X[,1], X[,2], med_xy)
    cl <- assigned$cluster
    d  <- assigned$dist

    max_rad <- max(d)
    best <- list(
      cluster = as.integer(cl),
      dist_to_medoid = as.numeric(d),
      k = k,
      max_cluster_radius = max_rad,
      method = "clara_radius",
      params = list(radius_m = radius_m, k_start = k_start, k_max = k_max,
                    samples = samples, sampsize = sampsize, seed = seed)
    )

    if (max_rad <= radius_m) break

    # optionally increase sampsize as k grows
    sampsize <- min(n, max(sampsize, 10L * k))
  }

  best
}
```

# engine helpers
```{r}
# Euclidean distance in meters between points
dist2 <- function(x1, y1, x2, y2) sqrt((x1 - x2)^2 + (y1 - y2)^2)

# Given a set of indices, choose a medoid by minimizing sum of distances (L1-ish)
choose_medoid <- function(x, y, idx) {
  if (length(idx) == 1) return(idx[1])
  xx <- x[idx]; yy <- y[idx]
  dx <- outer(xx, xx, "-")
  dy <- outer(yy, yy, "-")
  D  <- sqrt(dx^2 + dy^2)
  idx[which.min(rowSums(D))]
}

# For a set of medoid coordinates (k x 2), compute nearest medoid assignment + dist
assign_to_medoids <- function(x, y, medoids_xy) {
  k <- nrow(medoids_xy)
  if (k == 1) {
    d <- dist2(x, y, medoids_xy[1,1], medoids_xy[1,2])
    return(list(cluster = rep(1L, length(x)), dist = d))
  }
  # distance to each medoid (n x k); OK for k ~ tens/hundreds
  D <- sapply(seq_len(k), function(j) dist2(x, y, medoids_xy[j,1], medoids_xy[j,2]))
  cl <- max.col(-D)              # argmin
  d  <- D[cbind(seq_along(x), cl)]
  list(cluster = as.integer(cl), dist = as.numeric(d))
}

# Standardize engine output to a consistent structure.
# - Fills missing fields with defaults
# - Validates lengths/types
# - Coerces cluster to integer and normalizes NAs -> 0 (unassigned)
standardize_engine_output <- function(res, n, method = NA_character_, params = list()) {
  if (is.null(res) || !is.list(res)) {
    stop("Engine must return a list.")
  }
  if (is.null(res$cluster)) {
    stop("Engine output is missing `$cluster`.")
  }
  if (length(res$cluster) != n) {
    stop("Engine `$cluster` length (", length(res$cluster), ") != n (", n, ").")
  }

  # cluster: integer vector, with 0 reserved for unassigned
  cluster <- res$cluster
  if (is.factor(cluster)) cluster <- as.character(cluster)
  cluster <- suppressWarnings(as.integer(cluster))
  cluster[is.na(cluster)] <- 0L
  if (any(cluster < 0L)) stop("Engine `$cluster` must be >= 0 (0 = unassigned).")

  # dist_to_medoid: numeric length n or all NA
  dist_to_medoid <- res$dist_to_medoid
  if (is.null(dist_to_medoid)) {
    dist_to_medoid <- rep(NA_real_, n)
  } else {
    if (length(dist_to_medoid) != n) {
      stop("Engine `$dist_to_medoid` length (", length(dist_to_medoid), ") != n (", n, ").")
    }
    dist_to_medoid <- suppressWarnings(as.numeric(dist_to_medoid))
  }

  # k: integer scalar or NA
  k <- res$k
  if (is.null(k)) {
    k <- NA_integer_
  } else {
    if (length(k) != 1L) stop("Engine `$k` must be a scalar.")
    k <- suppressWarnings(as.integer(k))
  }

  # max_cluster_radius: numeric scalar or NA
  max_cluster_radius <- res$max_cluster_radius
  if (is.null(max_cluster_radius)) {
    max_cluster_radius <- NA_real_
  } else {
    if (length(max_cluster_radius) != 1L) stop("Engine `$max_cluster_radius` must be a scalar.")
    max_cluster_radius <- suppressWarnings(as.numeric(max_cluster_radius))
  }

  # method: string
  out_method <- res$method
  if (is.null(out_method) || is.na(out_method) || !nzchar(out_method)) {
    out_method <- method
  }
  if (is.null(out_method) || is.na(out_method) || !nzchar(out_method)) {
    out_method <- "unknown_engine"
  }

  # params: prefer engine's own params if supplied; else use provided
  out_params <- res$params
  if (is.null(out_params) || !is.list(out_params)) out_params <- params

  # If k missing, infer from cluster labels (excluding 0)
  if (is.na(k)) {
    k <- length(setdiff(unique(cluster), 0L))
  }

  # If max_cluster_radius missing but dist_to_medoid exists, infer
  if (is.na(max_cluster_radius) && any(!is.na(dist_to_medoid))) {
    max_cluster_radius <- suppressWarnings(max(dist_to_medoid, na.rm = TRUE))
    if (!is.finite(max_cluster_radius)) max_cluster_radius <- NA_real_
  }

  list(
    cluster = cluster,
    dist_to_medoid = dist_to_medoid,
    k = k,
    max_cluster_radius = max_cluster_radius,
    method = out_method,
    params = out_params
  )
}

evaluate_radius_pre_post <- function(sites_tbl,
                                     radius_m,
                                     engine_fun,
                                     method_name = "method",
                                     engine_params = list(),
                                     radius_param = "radius_m",
                                     year_single_col = "years_single") {
  stopifnot(all(c("AEZ","x","y","n_years") %in% names(sites_tbl)))
  stopifnot(year_single_col %in% names(sites_tbl))

  # ---- run engine (inject radius into params) ----
  params <- engine_params
  params[[radius_param]] <- radius_m

  raw <- do.call(engine_fun, c(list(df = sites_tbl), params))
  std <- standardize_engine_output(raw, n = nrow(sites_tbl), method = method_name, params = params)

  tmp <- sites_tbl
  tmp$cluster_within_aez <- std$cluster
  tmp$dist_to_medoid     <- std$dist_to_medoid

  # Helper: cluster-level table with size + estimated distinct years
  cluster_years_summary <- function(df_sites) {
    df_sites |>
      dplyr::filter(cluster_within_aez > 0L) |>
      dplyr::group_by(AEZ, cluster_within_aez) |>
      dplyr::summarise(
        n_sites = dplyr::n(),
        has_multi_year_site = any(n_years > 1, na.rm = TRUE),
        distinct_single_years = dplyr::n_distinct(df_sites[[year_single_col]][n_years == 1], na.rm = TRUE),
        est_distinct_years =
          ifelse(has_multi_year_site,
                 pmax(2, distinct_single_years),
                 distinct_single_years),
        max_dist_to_medoid = {
          m <- suppressWarnings(max(dist_to_medoid, na.rm = TRUE))
          if (is.finite(m)) m else NA_real_
        },
        .groups = "drop"
      )
  }

  pct_sites_in_2plus <- function(df_sites, cl_tbl) {
    if (nrow(cl_tbl) == 0) return(0)
    df_sites |>
      dplyr::left_join(
        cl_tbl |>
          dplyr::mutate(cluster_2plus = est_distinct_years >= 2L) |>
          dplyr::select(AEZ, cluster_within_aez, cluster_2plus),
        by = c("AEZ","cluster_within_aez")
      ) |>
      dplyr::mutate(cluster_2plus = dplyr::coalesce(cluster_2plus, FALSE)) |>
      dplyr::summarise(pct = 100 * mean(cluster_2plus)) |>
      dplyr::pull(pct)
  }

  stage_metrics <- function(df_sites, stage_label) {
    cl <- cluster_years_summary(df_sites)

    tibble::tibble(
      method = method_name,
      radius_km = radius_m / 1000,
      stage = stage_label,

      pct_clustered = 100 * mean(df_sites$cluster_within_aez > 0L),
      n_clusters = nrow(cl),
      n_aez_with_clusters = dplyr::n_distinct(df_sites$AEZ[df_sites$cluster_within_aez > 0L]),

      median_years_per_cluster = if (nrow(cl) == 0) NA_real_ else median(cl$est_distinct_years),
      pct_clusters_2plus_years = if (nrow(cl) == 0) NA_real_ else 100 * mean(cl$est_distinct_years >= 2L),

      pct_sites_in_2plus_year_clusters = pct_sites_in_2plus(df_sites, cl),

      median_cluster_size = if (nrow(cl) == 0) NA_real_ else median(cl$n_sites),
      max_radius_observed = if (nrow(cl) == 0) NA_real_ else max(cl$max_dist_to_medoid, na.rm = TRUE),

      effective_cluster_years = if (nrow(cl) == 0) NA_real_ else sum(cl$est_distinct_years, na.rm = TRUE)
    )
  }

  pre_row  <- stage_metrics(tmp, "pre_screen")
  tmp_post <- enforce_year_rule(tmp)
  post_row <- stage_metrics(tmp_post, "post_screen")

  dplyr::bind_rows(pre_row, post_row)
}
```

# AEZ wrappers
```{r}
# Create a simple logger closure that writes timestamped lines to a file.
# Usage:
#   logf <- make_logger("path/to/progress.log")
#   logf("INFO", "Starting AEZ ...")
make_logger <- function(log_path) {
  force(log_path)
  dir.create(dirname(log_path), showWarnings = FALSE, recursive = TRUE)
  function(level = "INFO", msg) {
    ts <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
    line <- sprintf("%s [%s] %s\n", ts, level, msg)
    cat(line, file = log_path, append = TRUE)
    invisible(NULL)
  }
}

# Build a stable checkpoint filepath for a given AEZ + method tag.
checkpoint_path <- function(checkpoint_dir, aez, method_tag) {
  file.path(checkpoint_dir, sprintf("aez=%s__method=%s.rds", as.character(aez), method_tag))
}

# Summarize cluster-level metadata from a sites table (unique-site level).
# Assumes sites_tbl has: AEZ, cluster_within_aez, dist_to_medoid (optional), x,y
summarize_clusters <- function(sites_tbl, method_tag) {
  stopifnot(all(c("AEZ", "cluster_within_aez") %in% names(sites_tbl)))

  has_dist <- "dist_to_medoid" %in% names(sites_tbl)

  clusters <- sites_tbl |>
    dplyr::filter(cluster_within_aez > 0L) |>
    dplyr::group_by(AEZ, cluster_within_aez) |>
    dplyr::summarise(
      n_sites = dplyr::n(),
      max_dist_to_medoid = if (has_dist) {
        m <- suppressWarnings(max(dist_to_medoid, na.rm = TRUE))
        if (is.finite(m)) m else NA_real_
      } else {
        NA_real_
      },
      .groups = "drop"
    ) |>
    dplyr::mutate(
      method = method_tag,
      cluster_id = paste0(AEZ, "_", method_tag, "_", cluster_within_aez)
    )

  clusters
}



# Run one AEZ through an engine, returning standardized sites-level output + engine meta.
# This is the "unit of checkpointing".
cluster_one_aez <- function(df_aez,
                            engine_fun,
                            method_tag,
                            engine_params = list(),
                            min_n = 2L,
                            switch_n_threshold = NULL,
                            fallback_engine_fun = NULL,
                            fallback_params = list(),
                            logf = NULL) {

  stopifnot(is.data.frame(df_aez))
  n <- nrow(df_aez)

  if (!all(c("x","y") %in% names(df_aez))) {
    stop("df_aez must contain columns x and y.")
  }

  # Minimal logger (no-op by default)
  if (is.null(logf)) logf <- function(level, msg) invisible(NULL)

  if (n < min_n) {
    df_aez$cluster_within_aez <- 0L
    df_aez$dist_to_medoid <- NA_real_

    return(list(
      sites = df_aez,
      engine_meta = list(
        method_tag = method_tag,              # run-level tag (stable)
        engine_method_used = NA_character_,
        params_used = engine_params,
        fallback_used = FALSE,
        fallback_reason = NA_character_,
        n_sites = n,
        k = 0L,
        max_cluster_radius = 0
      )
    ))
  }

  # ---- automatic fallback rule (engine selection happens here) ----
  use_engine_fun <- engine_fun
  use_params     <- engine_params
  fallback_used  <- FALSE
  fallback_reason <- NA_character_

  if (!is.null(switch_n_threshold) &&
      is.finite(switch_n_threshold) &&
      n > switch_n_threshold &&
      !is.null(fallback_engine_fun)) {

    use_engine_fun <- fallback_engine_fun
    use_params     <- fallback_params
    fallback_used  <- TRUE
    fallback_reason <- paste0("n_aez=", n, " > switch_n_threshold=", switch_n_threshold)

    logf("INFO", paste0("[FALLB] switching to fallback engine for AEZ n=", n,
                        " (", fallback_reason, ")"))
  }

  raw <- do.call(use_engine_fun, c(list(df = df_aez), use_params))

  # NOTE: `method=` here is *engine name hint* only; we keep run-level method_tag stable elsewhere
  std <- standardize_engine_output(
    raw,
    n = n,
    method = method_tag,
    params = use_params
  )

  df_aez$cluster_within_aez <- std$cluster
  df_aez$dist_to_medoid     <- std$dist_to_medoid

  list(
    sites = df_aez,
    engine_meta = list(
      method_tag = method_tag,              # run-level tag (stable)
      engine_method_used = std$method,       # actual engine label returned (e.g., "pam_radius" or "greedy_cover")
      params_used = std$params,              # params actually used (primary or fallback)
      fallback_used = fallback_used,
      fallback_reason = fallback_reason,
      n_sites = n,
      k = std$k,
      max_cluster_radius = std$max_cluster_radius
    )
  )
}

# Main wrapper: loop over AEZs with logging + per-AEZ checkpointing.
# Returns a list: sites (bound), clusters (metadata), run_meta (params)
cluster_within_aez <- function(sites_tbl,
                               engine_fun,
                               method_tag,
                               engine_params = list(),
                               min_n = 2L,
                               aez_order = NULL,
                               checkpoint_dir,
                               log_file = "progress.log",
                               fail_fast = TRUE,
                               switch_n_threshold = NULL,
                               fallback_engine_fun = NULL,
                               fallback_params = list()) {

  stopifnot(is.data.frame(sites_tbl))
  stopifnot("AEZ" %in% names(sites_tbl))
  dir.create(checkpoint_dir, showWarnings = FALSE, recursive = TRUE)

  log_path <- file.path(checkpoint_dir, log_file)
  logf <- make_logger(log_path)

  # Stable AEZ ordering
  if (is.null(aez_order)) {
    aez_order <- unique(as.character(sites_tbl$AEZ))
  }

  sites_tbl <- sites_tbl |>
    dplyr::mutate(AEZ = factor(AEZ, levels = aez_order)) |>
    dplyr::arrange(AEZ)

  aez_list <- split(sites_tbl, sites_tbl$AEZ, drop = TRUE)

  out_sites <- vector("list", length(aez_list))
  out_meta  <- vector("list", length(aez_list))
  names(out_sites) <- names(aez_list)
  names(out_meta)  <- names(aez_list)

  for (aez in names(aez_list)) {
    ckpt <- checkpoint_path(checkpoint_dir, aez, method_tag)

    if (file.exists(ckpt)) {
      logf("INFO", paste0("[SKIP ] ", aez, " (checkpoint exists: ", basename(ckpt), ")"))
      obj <- readRDS(ckpt)
      out_sites[[aez]] <- obj$sites
      out_meta[[aez]]  <- obj$engine_meta
      next
    }

    df_aez <- aez_list[[aez]]
    logf("INFO", paste0("[START] ", aez, " n=", nrow(df_aez)))

    obj <- tryCatch(
      {
        res <- cluster_one_aez(
          df_aez = df_aez,
          engine_fun = engine_fun,
          method_tag = method_tag,
          engine_params = engine_params,
          min_n = min_n,
          switch_n_threshold = switch_n_threshold,
          fallback_engine_fun = fallback_engine_fun,
          fallback_params = fallback_params,
          logf = logf
        )

        saveRDS(res, ckpt)
        logf("INFO", paste0("[DONE ] ", aez, " -> ", basename(ckpt),
                            " | engine_used=", res$engine_meta$engine_method_used,
                            if (isTRUE(res$engine_meta$fallback_used)) " (fallback)" else ""))

        res
      },
      error = function(e) {
        logf("ERROR", paste0("[FAIL ] ", aez, ": ", conditionMessage(e)))
        if (fail_fast) stop(e)

        df_aez$cluster_within_aez <- 0L
        df_aez$dist_to_medoid <- NA_real_

        list(
          sites = df_aez,
          engine_meta = list(
            method_tag = method_tag,
            engine_method_used = NA_character_,
            params_used = engine_params,
            fallback_used = FALSE,
            fallback_reason = NA_character_,
            n_sites = nrow(df_aez),
            k = NA_integer_,
            max_cluster_radius = NA_real_
          )
        )
      }
    )

    out_sites[[aez]] <- obj$sites
    out_meta[[aez]]  <- obj$engine_meta
  }

  sites_out <- dplyr::bind_rows(out_sites)

  # cluster-level metadata table (computed from sites_out)
  # NOTE: method_tag is the run tag; engine_used varies per AEZ and is stored in engine_meta_by_aez
  clusters_out <- summarize_clusters(sites_out, method_tag = method_tag)

  run_meta <- list(
    method_tag = method_tag,
    engine_params = engine_params,
    checkpoint_dir = checkpoint_dir,
    log_path = log_path,
    aez_order = aez_order,
    switch_n_threshold = switch_n_threshold,
    fallback_engine = if (!is.null(fallback_engine_fun)) deparse(substitute(fallback_engine_fun)) else NULL,
    fallback_params = fallback_params
  )

  list(
    sites = sites_out,
    clusters = clusters_out,
    engine_meta_by_aez = out_meta,
    run_meta = run_meta
  )
}
```

# Post-processing and export functions
```{r}
# Year rule enforcement at unique-site level.
# A cluster is valid if:
#   - at least one site has n_years > 1, OR
#   - cluster contains >= 2 distinct single-year values across its sites
#
# Expects sites_tbl has: AEZ, cluster_within_aez, n_years, years_single
# where years_single is either NA (if n_years > 1) or an integer year (if n_years == 1).
#
# Returns a modified sites_tbl with:
#   - year_rule_ok (logical)
#   - cluster_within_aez set to 0 for invalid clusters
enforce_year_rule <- function(sites_tbl) {
  req <- c("AEZ", "cluster_within_aez", "n_years", "years_single")
  missing <- setdiff(req, names(sites_tbl))
  if (length(missing) > 0) stop("sites_tbl missing required columns: ", paste(missing, collapse = ", "))

  # Build cluster validity flags
  validity <- sites_tbl |>
    dplyr::filter(cluster_within_aez > 0L) |>
    dplyr::group_by(AEZ, cluster_within_aez) |>
    dplyr::summarise(
      has_multi_year_site = any(n_years > 1, na.rm = TRUE),
      n_distinct_single_years = dplyr::n_distinct(years_single[n_years == 1], na.rm = TRUE),
      year_rule_ok = has_multi_year_site | (n_distinct_single_years >= 2L),
      .groups = "drop"
    ) |>
    dplyr::select(AEZ, cluster_within_aez, year_rule_ok)

  out <- sites_tbl |>
    dplyr::left_join(validity, by = c("AEZ","cluster_within_aez")) |>
    dplyr::mutate(
      year_rule_ok = dplyr::if_else(cluster_within_aez == 0L, TRUE, dplyr::coalesce(year_rule_ok, FALSE)),
      cluster_within_aez = dplyr::if_else(year_rule_ok, cluster_within_aez, 0L)
    )

  out
}

# Attach final identifiers after all post-processing (pure).
# Creates:
# - cluster_id string
# - cluster_missing flag (includes unassigned 0)
finalize_cluster_ids <- function(sites_tbl, method_tag) {
  stopifnot(all(c("AEZ","cluster_within_aez") %in% names(sites_tbl)))
  sites_tbl |>
    dplyr::mutate(
      method = method_tag,
      cluster_id = dplyr::if_else(
        cluster_within_aez > 0L,
        paste0(AEZ, "_", method_tag, "_", cluster_within_aez),
        NA_character_
      ),
      cluster_missing = (cluster_within_aez == 0L)
    )
}


# Join cluster labels from unique sites back onto the full time-series table (pure).
# join_by should match your site keys (AEZ, lat_r, lon_r).
join_clusters_back <- function(model_df_tagged,
                               sites_tbl_clustered,
                               join_by = c("AEZ","lat_r","lon_r")) {
  stopifnot(is.data.frame(model_df_tagged), is.data.frame(sites_tbl_clustered))
  stopifnot(all(join_by %in% names(model_df_tagged)))
  stopifnot(all(join_by %in% names(sites_tbl_clustered)))

  keep_cols <- c(join_by, "cluster_within_aez", "cluster_id", "method", "cluster_missing")
  if ("dist_to_medoid" %in% names(sites_tbl_clustered)) keep_cols <- c(keep_cols, "dist_to_medoid")
  if ("year_rule_ok" %in% names(sites_tbl_clustered))  keep_cols <- c(keep_cols, "year_rule_ok")

  rhs <- sites_tbl_clustered |>
    dplyr::select(dplyr::all_of(unique(keep_cols)))

  model_df_tagged |>
    dplyr::left_join(rhs, by = join_by) |>
    dplyr::mutate(cluster_missing = dplyr::coalesce(cluster_missing, TRUE))
}


# Export bundle (I/O only).
# Writes:
#  - RDS bundle
#  - CSV clustered full dataset
#  - CSV clustered unique sites
#  - CSV cluster metadata
export_bundle <- function(out_prefix,
                          model_df_clustered,
                          sites_tbl_clustered,
                          clusters_meta,
                          params = list()) {
  dir.create(dirname(out_prefix), showWarnings = FALSE, recursive = TRUE)

  bundle <- list(
    model_df_clustered = model_df_clustered,
    sites_tbl_clustered = sites_tbl_clustered,
    clusters_meta = clusters_meta,
    params = params
  )

  saveRDS(bundle, paste0(out_prefix, "_bundle.rds"))
  readr::write_csv(model_df_clustered, paste0(out_prefix, "_model_df_clustered.csv"))
  readr::write_csv(sites_tbl_clustered, paste0(out_prefix, "_sites_tbl_clustered.csv"))
  readr::write_csv(clusters_meta, paste0(out_prefix, "_clusters_meta.csv"))

  invisible(bundle)
}
```

## ============================================================
## PAM within each AEZ
## ============================================================
```{r}
MAX_R <- 12500 # 12500 meter radius, so maximum distance between sites is 25km
R_KM <- MAX_R / 1000
R_KM_STR <- formatC(R_KM, format = "f", digits = 1)


# Order smallest AEZs first (debug-friendly and makes early checkpoints fast)
aez_order <- sites_tbl |>
  dplyr::count(AEZ, name = "n_sites") |>
  dplyr::arrange(n_sites) |>
  dplyr::pull(AEZ) |>
  as.character()

# ---- Run clustering with caching + logging (wrapper owns it) ----
# Policy: PAM for AEZs with n <= 2000; greedy_cover fallback otherwise.
pam_run <- cluster_within_aez(
  sites_tbl      = sites_tbl,

  # primary engine
  engine_fun     = engine_pam_radius,
  method_tag     = "PAMR",
  engine_params  = list(
    radius_m = MAX_R,
    k_start  = 2L,
    k_max    = 200L
  ),

  # workflow controls
  min_n          = 2L,
  aez_order      = aez_order,
  checkpoint_dir = file.path(checkpoint_root, paste0("pam_rad_", R_KM_STR, "km")),
  log_file       = "progress.log",
  fail_fast      = TRUE,

  # fallback controls (no method_tag change; engine_meta records actual engine used)
  switch_n_threshold  = 1500L,
  fallback_engine_fun = engine_greedy_cover,
  fallback_params     = list(radius_m = MAX_R)
)

stopifnot("dist_to_medoid" %in% names(pam_run$sites))

# ---- Post-processing: year rule enforcement on unique sites ----
pam_sites_year <- pam_run$sites |>
  enforce_year_rule()

# Final IDs and flags (cluster_missing etc.)
pam_sites_final <- pam_sites_year |>
  finalize_cluster_ids(method_tag = "PAMR")

# Cluster metadata should be recomputed AFTER year-rule changes
pam_clusters_final <- summarize_clusters(pam_sites_final, method_tag = "PAMR")

# ---- Join back to the full time-series dataset ----
model_df_pam <- join_clusters_back(
  model_df_tagged     = model_df_tagged,
  sites_tbl_clustered = pam_sites_final,
  join_by             = c("AEZ", "lat_r", "lon_r")
)

# ---- Export bundle (single call; all I/O here) ----
pam_bundle <- export_bundle(
  out_prefix = file.path(out_dir, paste0("pam_rad_", R_KM_STR, "km")),
  model_df_clustered  = model_df_pam,
  sites_tbl_clustered = pam_sites_final,
  clusters_meta       = pam_clusters_final,
  params = list(
    run_tag = "PAMR",
    primary_engine = "engine_pam_radius",
    fallback_engine = "engine_greedy_cover",
    switch_n_threshold = 1500L,
    radius_m = MAX_R,
    engine_params = pam_run$run_meta$engine_params,
    fallback_params = pam_run$run_meta$fallback_params,
    checkpoint_dir = pam_run$run_meta$checkpoint_dir,
    log_path = pam_run$run_meta$log_path
  )
)

# Quick diagnostics
summary(pam_clusters_final$max_dist_to_medoid)
table(sapply(pam_run$engine_meta_by_aez, `[[`, "engine_method_used"), useNA = "ifany")
```

## ============================================================
## Greedy Cover within each AEZ
## ============================================================
```{r}
MAX_R <- 12500 # meters, maximum site distance is twice this
R_KM <- MAX_R / 1000
R_KM_STR <- formatC(R_KM, format = "f", digits = 1)

# Order smallest AEZs first (debug-friendly and makes early checkpoints fast)
aez_order <- sites_tbl |>
  dplyr::count(AEZ, name = "n_sites") |>
  dplyr::arrange(n_sites) |>
  dplyr::pull(AEZ) |>
  as.character()

# ---- Run clustering with caching + logging (wrapper owns it) ----
greedy_run <- cluster_within_aez(
  sites_tbl      = sites_tbl,
  engine_fun     = engine_greedy_cover,
  method_tag     = "GRDY",
  engine_params  = list(radius_m = MAX_R),

  # workflow controls
  min_n          = 2L,
  aez_order      = aez_order,
  checkpoint_dir = file.path(checkpoint_root, paste0("greedy_cover_rad_", R_KM_STR, "km")),
  log_file       = "progress.log",
  fail_fast      = TRUE
)

stopifnot("dist_to_medoid" %in% names(greedy_run$sites))

# ---- Post-processing: year rule enforcement on unique sites ----
greedy_sites_year <- greedy_run$sites |>
  enforce_year_rule()

# Final IDs and flags (cluster_missing etc.)
greedy_sites_final <- greedy_sites_year |>
  finalize_cluster_ids(method_tag = "GRDY")

# Cluster metadata should be recomputed AFTER year-rule changes
greedy_clusters_final <- summarize_clusters(greedy_sites_final, method_tag = "GRDY")

# ---- Join back to the full time-series dataset ----
model_df_greedy <- join_clusters_back(
  model_df_tagged     = model_df_tagged,
  sites_tbl_clustered = greedy_sites_final,
  join_by             = c("AEZ", "lat_r", "lon_r")
)

# ---- Export bundle (single call; all I/O here) ----
greedy_bundle <- export_bundle(
  out_prefix = file.path(out_dir, paste0("greedy_cover_rad_", R_KM_STR, "km")),
  model_df_clustered  = model_df_greedy,
  sites_tbl_clustered = greedy_sites_final,
  clusters_meta       = greedy_clusters_final,
  params = list(
    run_tag = "GRDY",
    engine = "engine_greedy_cover",
    radius_m = MAX_R,
    engine_params = greedy_run$run_meta$engine_params,
    checkpoint_dir = greedy_run$run_meta$checkpoint_dir,
    log_path = greedy_run$run_meta$log_path
  )
)

# Quick diagnostics
summary(greedy_clusters_final$max_dist_to_medoid)

```

## probing different radius options for greedy_cover
```{r}
radius_grid_km <- seq(8, 50, by = 1)
radius_grid_m  <- radius_grid_km * 1000

tradeoff_both <- purrr::map_dfr(
  radius_grid_m,
  ~evaluate_radius_pre_post(
    sites_tbl = sites_tbl,
    radius_m = .x,
    engine_fun = engine_greedy_cover,
    method_name = "GREEDY",
    engine_params = list(),          # greedy only needs radius_m; helper injects it
    radius_param = "radius_m",
    year_single_col = "years_single" # change to "year_single" if that's your column
  )
)

tradeoff_pre  <- tradeoff_both |> dplyr::filter(stage == "pre_screen")
tradeoff_post <- tradeoff_both |> dplyr::filter(stage == "post_screen")

ggplot(tradeoff_both, aes(x = radius_km, y = pct_sites_in_2plus_year_clusters, linetype = stage)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Greedy: % of sites in ≥2-year clusters (pre vs post year-screen)",
    x = "Radius (km)",
    y = "% of unique sites"
  )

ggplot(tradeoff_post, aes(x = radius_km, y = median_cluster_size)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Median cluster size (post-screen)",
    x = "Radius (km)",
    y = "Median cluster size"
  )

ggplot(tradeoff_post, aes(x = radius_km, y = n_aez_with_clusters)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "AEZs with ≥1 usable cluster",
    x = "Radius (km)",
    y = "Number of AEZs"
  )

ggplot(tradeoff_post, aes(x = radius_km, y = effective_cluster_years)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Total cluster-year observations (post-screen)",
    x = "Radius (km)",
    y = "Sum of distinct years across clusters"
  )
```

## ============================================================
## CLARA within each AEZ
## - scalable medoid approximation (no full n^2 distance matrix)
## - still enforces radius by increasing k until max(dist_to_medoid) <= MAX_R
## ============================================================

```{r}
MAX_R <- 12500 # 12500 meter radius, so maximum distance between sites is 25km
R_KM <- MAX_R / 1000
R_KM_STR <- formatC(R_KM, format = "f", digits = 1)



# Order smallest AEZs first (debug-friendly and makes early checkpoints fast)
aez_order <- sites_tbl |>
  dplyr::count(AEZ, name = "n_sites") |>
  dplyr::arrange(n_sites) |>
  dplyr::pull(AEZ) |>
  as.character()

# ---- Run clustering with caching + logging (wrapper owns it) ----
clara_run <- cluster_within_aez(
  sites_tbl      = sites_tbl,
  engine_fun     = engine_clara,
  method_tag     = "CLARA",
  engine_params  = list(
    radius_m  = MAX_R,
    k_start   = 2L,
    k_max     = 500L,
    samples   = 5L,
    sampsize  = NULL,   # let engine pick a reasonable default and grow with k
    seed      = 1L
  ),

  # workflow controls
  min_n          = 2L,
  aez_order      = aez_order,
  checkpoint_dir = file.path(checkpoint_root, paste0("clara_rad_", R_KM_STR, "km")),
  log_file       = "progress.log",
  fail_fast      = TRUE,

  # optional safety net: if CLARA struggles on huge AEZs, fall back to greedy
  # (keep or drop; recommended to keep for robustness)
  switch_n_threshold  = 2000L,
  fallback_engine_fun = engine_greedy_cover,
  fallback_params     = list(radius_m = MAX_R)
)

stopifnot("dist_to_medoid" %in% names(clara_run$sites))

# ---- Post-processing: year rule enforcement on unique sites ----
clara_sites_year <- clara_run$sites |>
  enforce_year_rule()

# Final IDs and flags (cluster_missing etc.)
clara_sites_final <- clara_sites_year |>
  finalize_cluster_ids(method_tag = "CLARA")

# Cluster metadata should be recomputed AFTER year-rule changes
clara_clusters_final <- summarize_clusters(clara_sites_final, method_tag = "CLARA")

# ---- Join back to the full time-series dataset ----
model_df_clara <- join_clusters_back(
  model_df_tagged     = model_df_tagged,
  sites_tbl_clustered = clara_sites_final,
  join_by             = c("AEZ", "lat_r", "lon_r")
)

# ---- Export bundle (single call; all I/O here) ----
clara_bundle <- export_bundle(
  out_prefix = file.path(out_dir, paste0("clara_rad_", R_KM_STR, "km")),
  model_df_clustered  = model_df_clara,
  sites_tbl_clustered = clara_sites_final,
  clusters_meta       = clara_clusters_final,
  params = list(
    run_tag = "CLARA",
    engine = "engine_clara",
    radius_m = MAX_R,
    engine_params = clara_run$run_meta$engine_params,
    switch_n_threshold = clara_run$run_meta$switch_n_threshold,
    fallback_engine = clara_run$run_meta$fallback_engine,
    fallback_params = clara_run$run_meta$fallback_params,
    checkpoint_dir = clara_run$run_meta$checkpoint_dir,
    log_path = clara_run$run_meta$log_path
  )
)

# Quick diagnostics
summary(clara_clusters_final$max_dist_to_medoid)
table(sapply(clara_run$engine_meta_by_aez, `[[`, "engine_method_used"), useNA = "ifany")

```

## CLARA radius sensitivity
```{r}
radius_grid_km <- seq(8, 25, by = 1)
radius_grid_m  <- radius_grid_km * 1000

clara_tradeoff_both <- purrr::map_dfr(
  radius_grid_m,
  ~evaluate_radius_pre_post(
    sites_tbl   = sites_tbl,
    radius_m    = .x,
    engine_fun  = engine_clara,
    method_name = "CLARA",
    engine_params = list(
      k_start  = 2L,
      k_max    = 800L,   # if you see max_radius_observed > radius target, raise this
      samples  = 5L,
      sampsize = NULL,   # let engine decide; or set e.g. 2000L for stability
      seed     = 1L
    ),
    radius_param    = "radius_m",
    year_single_col = "years_single"
  )
)

clara_tradeoff_pre  <- clara_tradeoff_both |> dplyr::filter(stage == "pre_screen")
clara_tradeoff_post <- clara_tradeoff_both |> dplyr::filter(stage == "post_screen")

ggplot(clara_tradeoff_both, aes(x = radius_km, y = pct_sites_in_2plus_year_clusters, linetype = stage)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "CLARA: % of sites in ≥2-year clusters (pre vs post year-screen)",
    x = "Radius (km)",
    y = "% of unique sites"
  )

ggplot(clara_tradeoff_post, aes(x = radius_km, y = median_cluster_size)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "CLARA: Median cluster size (post-screen)",
    x = "Radius (km)",
    y = "Median cluster size"
  )

ggplot(clara_tradeoff_post, aes(x = radius_km, y = n_aez_with_clusters)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "CLARA: AEZs with ≥1 usable cluster (post-screen)",
    x = "Radius (km)",
    y = "Number of AEZs"
  )

ggplot(clara_tradeoff_post, aes(x = radius_km, y = effective_cluster_years)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "CLARA: Total cluster-year observations (post-screen)",
    x = "Radius (km)",
    y = "Sum of distinct years across clusters"
  )

# Optional but recommended sanity check: is CLARA actually meeting the radius cap?
ggplot(clara_tradeoff_pre, aes(x = radius_km, y = max_radius_observed / 1000)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "CLARA: Observed max dist-to-medoid vs target (pre-screen)",
    x = "Target radius (km)",
    y = "Observed max dist-to-medoid (km)"
  )
```

## ============================================================
## Diagnostics (works for PAM / greedy cover / CLARA bundles)
## ============================================================
```{r}
# helpers: parse radius + method from filename
parse_radius_km <- function(path) {
  x <- basename(path)
  m <- str_match(x, "([0-9]+(?:\\.[0-9]+)?)km")[, 2]
  as.numeric(m)
}

parse_method <- function(path) {
  x <- tolower(basename(path))
  if (str_detect(x, "clara"))  return("CLARA")
  if (str_detect(x, "pam"))    return("PAM")
  if (str_detect(x, "greedy")) return("GREEDY")
  "UNKNOWN"
}

# ---- compute cluster-year estimate from your site-level fields ----
cluster_years_summary <- function(sites, year_single_col = "years_single") {
  stopifnot(all(c("AEZ", "cluster_within_aez", "n_years") %in% names(sites)))
  stopifnot(year_single_col %in% names(sites))

  sites |>
    filter(cluster_within_aez > 0L) |>
    group_by(AEZ, cluster_within_aez) |>
    summarise(
      n_sites = n(),
      has_multi_year_site = any(n_years > 1, na.rm = TRUE),
      distinct_single_years = n_distinct(sites[[year_single_col]][n_years == 1], na.rm = TRUE),
      est_distinct_years = ifelse(has_multi_year_site, pmax(2, distinct_single_years), distinct_single_years),
      max_dist_to_medoid = if ("dist_to_medoid" %in% names(sites)) {
        m <- suppressWarnings(max(dist_to_medoid, na.rm = TRUE))
        if (is.finite(m)) m else NA_real_
      } else NA_real_,
      .groups = "drop"
    )
}

# ---- one bundle -> one row of metrics ----
summarize_bundle <- function(bundle, method, radius_km, year_single_col = "years_single") {
  sites <- bundle$sites_tbl_clustered
  stopifnot(!is.null(sites))

  # if cluster_missing not present, derive it
  if (!("cluster_missing" %in% names(sites))) {
    sites <- sites |>
      mutate(cluster_missing = cluster_within_aez == 0L | is.na(cluster_within_aez))
  }

  n_total_sites <- nrow(sites)
  n_clustered_sites <- sum(sites$cluster_within_aez > 0L, na.rm = TRUE)

  cl <- cluster_years_summary(sites, year_single_col = year_single_col)

  tibble(
    method = method,
    radius_km = radius_km,

    n_sites_total = n_total_sites,
    pct_sites_clustered = 100 * n_clustered_sites / n_total_sites,

    n_clusters = nrow(cl),
    n_clusters_2plus_years = sum(cl$est_distinct_years >= 2, na.rm = TRUE),
    pct_clusters_2plus_years = if (nrow(cl) == 0) NA_real_ else 100 * mean(cl$est_distinct_years >= 2, na.rm = TRUE),

    n_aez_with_clusters = n_distinct(sites$AEZ[sites$cluster_within_aez > 0L]),
    effective_cluster_years = if (nrow(cl) == 0) 0 else sum(cl$est_distinct_years, na.rm = TRUE),

    median_cluster_size = if (nrow(cl) == 0) NA_real_ else median(cl$n_sites, na.rm = TRUE),
    p90_cluster_size = if (nrow(cl) == 0) NA_real_ else as.numeric(quantile(cl$n_sites, 0.9, na.rm = TRUE)),
    max_cluster_size = if (nrow(cl) == 0) NA_real_ else max(cl$n_sites, na.rm = TRUE),

    max_radius_observed_km = if (nrow(cl) == 0) NA_real_ else max(cl$max_dist_to_medoid, na.rm = TRUE) / 1000
  )
}

## ----------------------------
## Load bundles from ./outputs
## (relative to the Rmd location)
## ----------------------------
bundle_dir <- out_dir
bundle_paths <- list.files(bundle_dir, pattern = "_bundle\\.rds$", full.names = TRUE)

stopifnot(length(bundle_paths) > 0)

# radii
target_radii <- c(12.5, 20, 25)

bundle_tbl_raw <- map_dfr(bundle_paths, function(p) {
  rad <- parse_radius_km(p)
  meth <- parse_method(p)
  if (is.na(rad) || !(rad %in% target_radii)) return(NULL)
  b <- readRDS(p)
  summarize_bundle(bundle = b, method = meth, radius_km = rad, year_single_col = "years_single")
})

# build table of metrics
bundle_tbl <- map_dfr(bundle_paths, function(p) {
  rad <- parse_radius_km(p)
  meth <- parse_method(p)

  # skip files that don't parse cleanly
  if (is.na(rad) || !(rad %in% target_radii)) return(NULL)

  b <- readRDS(p)
  summarize_bundle(
    bundle = b,
    method = meth,
    radius_km = rad,
    year_single_col = "years_single"
  )
}) |>
  arrange(method, radius_km)

## ----------------------------
## Decision table (clean cols)
## ----------------------------
decision_tbl <- bundle_tbl |>
  arrange(method, radius_km) |>
  mutate(
    pct_sites_clustered = round(pct_sites_clustered, 1),
    median_cluster_size = round(median_cluster_size, 1),
    p90_cluster_size = round(p90_cluster_size, 1),
    max_radius_observed_km = round(max_radius_observed_km, 2)
  ) |>
  select(
    method,
    radius_km,
    pct_sites_clustered,
    n_clusters,
    n_aez_with_clusters,
    effective_cluster_years,
    median_cluster_size,
    p90_cluster_size,
    max_radius_observed_km
  )

decision_tbl
write_csv(decision_tbl, file.path(summary_dir, "cluster_method_comparison.csv"))


## ----------------------------
## Number of clusters in each AEZ by method
## ----------------------------

# get the full AEZ roster from the first bundle
get_all_aez <- function(bundle) {
  sort(unique(bundle$sites_tbl_clustered$AEZ))
}

clusters_by_aez_one_bundle <- function(bundle, method, radius_km) {
  sites <- bundle$sites_tbl_clustered
  stopifnot(!is.null(sites))
  stopifnot(all(c("AEZ","cluster_within_aez") %in% names(sites)))

  all_aez <- sort(unique(sites$AEZ))

  # count unique clusters within each AEZ (exclude 0/noise)
  counts <- sites |>
    filter(cluster_within_aez > 0L) |>
    mutate(cluster_key = paste(AEZ, cluster_within_aez, sep = "::")) |>
    group_by(AEZ) |>
    summarise(n_clusters = n_distinct(cluster_key), .groups = "drop")

  # ensure AEZs with 0 clusters appear
  tibble(AEZ = all_aez) |>
    left_join(counts, by = "AEZ") |>
    mutate(
      n_clusters = tidyr::replace_na(n_clusters, 0L),
      method = method,
      radius_km = radius_km
    ) |>
    select(method, radius_km, AEZ, n_clusters)
}

clusters_by_aez_tbl <- map_dfr(bundle_paths, function(p) {
  rad <- parse_radius_km(p)
  meth <- parse_method(p)

  if (is.na(rad) || !(rad %in% target_radii)) return(NULL)

  b <- readRDS(p)
  clusters_by_aez_one_bundle(b, meth, rad)
}) |>
  arrange(method, radius_km, AEZ)

# Wide view: AEZ rows, columns = method_radius
clusters_by_aez_wide <- clusters_by_aez_tbl |>
  mutate(method_radius = paste0(method, "_", radius_km, "km")) |>
  select(AEZ, method_radius, n_clusters) |>
  tidyr::pivot_wider(names_from = method_radius, values_from = n_clusters)


clusters_by_aez_tbl
write_csv(clusters_by_aez_tbl, file.path(summary_dir, "clusters_by_AEZ_long.csv"))
clusters_by_aez_wide
write_csv(clusters_by_aez_wide, file.path(summary_dir, "clusters_by_AEZ_wide.csv"))
```

# Build the overall PREDICTS dataset with cluster assignment across method/radius combination
```{r}
# Read AEZ polygons
if (!file.exists(aez_path)) {
  stop("AEZ shapefile not found at: ", normalizePath(aez_path, mustWork = FALSE))
}
aez <- read_sf(aez_path) %>%
  st_make_valid() %>%
  select(Id, AEZ)


# build comparison dataset, with each site assigned (1) its cluster from the 9 method x radius combinations and (2) its minmaxxed OV score relative to all sites
library(dplyr)
library(purrr)
library(readr)
library(stringr)

# parse radius + method from filenames ----
parse_radius_km <- function(path) {
  x <- basename(path)
  m <- str_match(x, "([0-9]+(?:\\.[0-9]+)?)km")[, 2]
  as.numeric(m)
}

parse_method <- function(path) {
  x <- tolower(basename(path))
  if (str_detect(x, "clara"))  return("CLARA")
  if (str_detect(x, "pam"))    return("PAM")
  if (str_detect(x, "greedy")) return("GREEDY")
  "UNKNOWN"
}

# locate the filess
paths <- list.files(out_dir, pattern = "_model_df_clustered\\.csv$", full.names = TRUE)
stopifnot(length(paths) > 0)

# read file and keep only needed columns. Then, build (sample_id -> cluster_id_<method>_<radius>) table per file ----
cluster_cols_list <- map(paths, function(p) {
  meth <- parse_method(p)
  rad  <- parse_radius_km(p)

  if (meth == "UNKNOWN" || is.na(rad)) return(NULL)

  cluster_col <- paste0("cluster_id_", meth, "_", rad, "km")

  dat <- read_csv(p, show_col_types = FALSE)

  # Keep only sample_id + cluster_id, rename cluster_id -> cluster_col
  out <- dat %>%
    select(sample_id, cluster_id) %>%
    rename(!!cluster_col := cluster_id)

  # sanity check: sample_id should be unique in each file
  if (any(duplicated(out$sample_id))) {
    dup_n <- sum(duplicated(out$sample_id))
    stop("File has duplicated sample_id (", dup_n, " dups): ", basename(p))
  }

  out
}) %>% purrr::compact()

# sanity check: do we have 9?
if (length(cluster_cols_list) != 9) {
  warning("Expected 9 clustering files (3 methods × 3 radii) but found ", length(cluster_cols_list),
          ". Check output/ and filename patterns.")
}

# ---- build the base table from the first file (these fields should be identical across runs) ----
base <- read_csv(paths[[1]], show_col_types = FALSE) %>%
  select(
    sample_id, AEZ, Latitude, Longitude, Sample_midpoint, ov_score,
    Id, lat_r, lon_r, year
  ) %>%
  distinct(sample_id, .keep_all = TRUE)

# to check, base must be unique per sample_id
if (any(duplicated(base$sample_id))) {
  stop("Base table still has duplicated sample_id after distinct(). That means sample_id is not unique.")
}

# ---- join all 9 cluster columns onto the base ----
clusters_wide <- reduce(cluster_cols_list, left_join, by = "sample_id")

master <- base %>%
  left_join(clusters_wide, by = "sample_id")

# ---- add globally min-maxed OV ----
rng <- range(master$ov_score, na.rm = TRUE)
master <- master %>%
  mutate(
    ov_minmax_global = if (rng[2] == rng[1]) NA_real_ else (ov_score - rng[1]) / (rng[2] - rng[1])
  )

# ---- write ----
write_csv(master, file.path(summary_dir, "predicts_method_radius_clusters.csv"))
```


## Calculate metrics across columns for use during comparison
```{r}
master <- read_csv(file.path(summary_dir, "predicts_method_radius_clusters.csv"), show_col_types = FALSE)

# Identify the 9 cluster columns
cluster_cols <- names(master)[str_detect(names(master), "^cluster_id_")]

# Reshape to long format: one row per sample_id per method/radius
# Long format
long <- master %>%
  pivot_longer(
    cols = all_of(cluster_cols),
    names_to = "cluster_key",
    values_to = "cluster_id"
  ) %>%
  mutate(
    method = str_match(cluster_key, "^cluster_id_([^_]+)_")[,2],
    radius_km = as.numeric(str_match(cluster_key, "_([0-9]+(?:\\.[0-9]+)?)km$")[,2])
  ) %>%
  filter(!is.na(method), !is.na(radius_km)) %>%
  filter(!is.na(cluster_id), cluster_id != 0) %>%
  filter(!is.na(AEZ),
         !is.na(ov_score),
         !is.na(ov_minmax_global))
```

## For each cluster, determine range between min and max OV. Then, determine largest range within each AEZ, and quantiles. Done for each of the 9 mmethod/radius combinations for clustering. 
```{r}
# calculate cluster-level ranges, p25, p75, IQR, for both raw and minmax OV
cluster_level_information <- long %>%
  group_by(method, radius_km, AEZ, cluster_id) %>%
  summarise(
    n_obs = n(),
    raw_min = min(ov_score, na.rm = TRUE),
    raw_max = max(ov_score, na.rm = TRUE),
    raw_range = raw_max - raw_min,
    raw_p25 = quantile(ov_score, 0.25, na.rm = TRUE),
    raw_p75 = quantile(ov_score, 0.75, na.rm = TRUE),
    raw_iqr = raw_p75-raw_p25,

    mm_min = min(ov_minmax_global, na.rm = TRUE),
    mm_max = max(ov_minmax_global, na.rm = TRUE),
    mm_range = mm_max - mm_min,
    mm_p25 = quantile(ov_minmax_global, 0.25, na.rm = TRUE),
    mm_p75 = quantile(ov_minmax_global, 0.75, na.rm = TRUE),
    mm_iqr = mm_p75-mm_p25,
    .groups = "drop"
  )

# Find AEZ-level information based on cluster information
aez_level_information <- cluster_level_information %>%
  group_by(method, radius_km, AEZ) %>%
  summarise(
    n_clusters_in_aez = n(),
    
    # maximum OV range, both raw and minmax, within a single cluster in AEZ
    max_ov_range_raw = max(raw_range, na.rm = TRUE),
    max_ov_range_minmax = max(mm_range, na.rm = TRUE),
    
    # Medians across clusters in this AEZ
    # median_cluster_raw_p25 = median(raw_p25, na.rm = TRUE),
    # median_cluster_raw_p75 = median(raw_p75, na.rm = TRUE),
    median_cluster_raw_iqr = median(raw_iqr, na.rm = TRUE),

    # median_cluster_mm_p25  = median(mm_p25, na.rm = TRUE),
    # median_cluster_mm_p75  = median(mm_p75, na.rm = TRUE),
    median_cluster_mm_iqr  = median(mm_iqr, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  arrange(method, radius_km, AEZ)

# write results
write_csv(aez_level_information,
          file.path(summary_dir, "aez_level_information.csv"))

# aez_max_range
# method_radius_summary
```

## Visualizations
```{r}

library(ggplot2)

# plot the distribtion across AEZs, for eachregime
ggplot(aez_level_information,
       aes(x = factor(radius_km),
           y = median_cluster_raw_iqr,
           fill = method)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    x = "Radius (km)",
    y = "Median Cluster IQR (Raw OV)",
    title = "Distribution Across AEZs of Typical Cluster Spread"
  )

method_summary <- aez_level_information %>%
  group_by(method, radius_km) %>%
  summarise(
    median_aez_median_iqr =
      median(median_cluster_raw_iqr, na.rm = TRUE),
    p25 =
      quantile(median_cluster_raw_iqr, 0.25, na.rm = TRUE),
    p75 =
      quantile(median_cluster_raw_iqr, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# median + IQR error bars
ggplot(method_summary,
       aes(x = radius_km,
           y = median_aez_median_iqr,
           color = method,
           group = method)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = p25, ymax = p75),
                width = 0.5) +
  labs(
    x = "Radius (km)",
    y = "Median AEZ-Level Median Cluster IQR",
    title = "Typical Cluster Spread by Method and Radius"
  )

library(dplyr)
library(ggplot2)

df20 <- aez_level_information %>%
  filter(radius_km == 12.5)

# Order AEZs by the raw max range (you can change to minmax or IQR if you prefer)
aez_order <- df20 %>%
  group_by(AEZ) %>%
  summarise(order_key = max(max_ov_range_raw, na.rm = TRUE), .groups = "drop") %>%
  arrange(order_key)

df20 <- df20 %>%
  mutate(AEZ = factor(AEZ, levels = aez_order$AEZ))

seg_for <- function(df, value_col) {
  df %>%
    group_by(AEZ) %>%
    summarise(
      lo = min(.data[[value_col]], na.rm = TRUE),
      hi = max(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    )
}

## max within-AEZ cluster OV range - raw
seg_raw <- seg_for(df20, "max_ov_range_raw")

ggplot(df20, aes(y = AEZ)) +
  geom_segment(data = seg_raw,
               aes(y = AEZ, yend = AEZ, x = lo, xend = hi),
               color = "grey80") +
  geom_point(aes(x = max_ov_range_raw, color = method), size = 2) +
  labs(
    title = "20km: Max within-AEZ cluster OV range (raw) by method",
    x = "Max OV range within AEZ (raw)",
    y = ""
  )

## max within-AEZ cluster OV range - minmax
seg_mm <- seg_for(df20, "max_ov_range_minmax")

ggplot(df20, aes(y = AEZ)) +
  geom_segment(data = seg_mm,
               aes(y = AEZ, yend = AEZ, x = lo, xend = hi),
               color = "grey80") +
  geom_point(aes(x = max_ov_range_minmax, color = method), size = 2) +
  labs(
    title = "20km: Max within-AEZ cluster OV range (minmax) by method",
    x = "Max OV range within AEZ (minmax)",
    y = ""
  )

## IQR diagnostic raw
seg_iqr_raw <- seg_for(df20, "median_cluster_raw_iqr")

ggplot(df20, aes(y = AEZ)) +
  geom_segment(data = seg_iqr_raw,
               aes(y = AEZ, yend = AEZ, x = lo, xend = hi),
               color = "grey80") +
  geom_point(aes(x = median_cluster_raw_iqr, color = method), size = 2) +
  labs(
    title = "20km: Typical within-cluster spread (median cluster IQR, raw) by method",
    x = "Median cluster IQR within AEZ (raw)",
    y = ""
  )

## IQR diagnostic minmax
seg_iqr_mm <- seg_for(df20, "median_cluster_mm_iqr")

ggplot(df20, aes(y = AEZ)) +
  geom_segment(data = seg_iqr_mm,
               aes(y = AEZ, yend = AEZ, x = lo, xend = hi),
               color = "grey80") +
  geom_point(aes(x = median_cluster_mm_iqr, color = method), size = 2) +
  labs(
    title = "20km: Typical within-cluster spread (median cluster IQR, minmax) by method",
    x = "Median cluster IQR within AEZ (minmax)",
    y = ""
  )


# 
# # dumbbell plott for raw ov score ranges
# df20_range <- df20 %>%
#   group_by(AEZ) %>%
#   summarise(lo = min(max_ov_range_raw, na.rm = TRUE),
#             hi = max(max_ov_range_raw, na.rm = TRUE),
#             .groups = "drop")
# 
# # order AEZs by typical magnitude (or by max)
# df20 <- df20 %>%
#   group_by(AEZ) %>%
#   mutate(aez_rank = max(max_ov_range_raw, na.rm = TRUE)) %>%
#   ungroup() %>%
#   mutate(AEZ = reorder(AEZ, aez_rank))
# 
# df20_range <- df20_range %>%
#   left_join(df20 %>% distinct(AEZ, aez_rank), by = "AEZ") %>%
#   mutate(AEZ = reorder(AEZ, aez_rank))
# 
# ggplot(df20, aes(y = AEZ)) +
#   geom_segment(data = df20_range,
#                aes(y = AEZ, yend = AEZ, x = lo, xend = hi),
#                color = "grey80") +
#   geom_point(aes(x = max_ov_range_raw, color = method), size = 2) +
#   labs(
#     title = "20km: Max within-AEZ cluster OV range (raw) by method",
#     x = "Max OV range within AEZ (raw)",
#     y = ""
#   )

# scatterplot of methods against each other to see if similar results regarding OV
wide20_raw <- df20 %>%
  select(AEZ, method, max_ov_range_raw) %>%
  pivot_wider(names_from = method, values_from = max_ov_range_raw)

# PAM vs CLARA
ggplot(wide20_raw, aes(x = PAM, y = CLARA)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "12.5km (raw): PAM vs CLARA AEZ max OV range",
    x = "PAM",
    y = "CLARA"
  )

# PAM vs GREEDY
ggplot(wide20_raw, aes(x = PAM, y = GREEDY)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "12.5km (raw): PAM vs GREEDY AEZ max OV range",
    x = "PAM",
    y = "GREEDY"
  )

# CLARA vs GREEDY
ggplot(wide20_raw, aes(x = CLARA, y = GREEDY)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "12.5km (raw): CLARA vs GREEDY AEZ max OV range",
    x = "CLARA",
    y = "GREEDY"
  )


```

