---
title: "Database Filtering and Taxonomic Classification"
output: html_notebook
# Note: edited by Eli Duggan in May 2025 to clarify and use the updated BioTIME 2.0 database. 
---

# Install taxonomic classification packages
```{r}
library(tidyverse)
install.packages(c("remotes", "taxize", "stringr", "dplyr"))
remotes::install_github("scottfjennings/birdnames")
```
# Load packages
```{r}
# library(birdnames)
library(taxize)
library(dplyr)
library(stringr)
```

# load BioTIME .rds and convert to usable file types
```{r}
# Read .rds, write to .csv, and save as .RData
# biotime_data_rds <- "biotime_v2_query_15April25.rds"
# biotime_data <- readRDS(biotime_data_rds)
# write.csv(biotime_data, "biotime_data.csv", row.names = FALSE)
# save(biotime_data, file = "biotime_data.RData")

load("biotime_data.RData") # if in a new session
```

# exploration of data
```{r}
head(biotime_data, 50)   # show top 50 rows
```
# Function to filter unusable data entries
```{r}
# remove missing entries, morphospecies, high resolutions (sp. optional)
filter_species <- function(df, species_column = "valid_name") {
  original_count <- nrow(df)
  
  remove_resolutions <- c("phylum", "subphylum", "superclass", "kingdom") # EDIT AS NEEDED
  
  df_clean <- df %>%
    filter(
      # remove missing names (0, 0%)
      !is.na(!!sym(species_column)),
      
      # remove morphospecies & unidentifiable (165, 0%)
      !str_detect(!!sym(species_column), regex("morphospecies|unidentif|unknown|undetermined", ignore_case = TRUE)),
      
      # remove rows with resolution too low (84,413, 0.7%)
      !tolower(resolution) %in% remove_resolutions,
      
      # Exclude sp., spp., sp1, etc. entries (179,601, 1.5%)?
      !str_detect(!!sym(species_column), regex("\\s(sp|spp)\\.?$", ignore_case = TRUE)),
      !str_detect(!!sym(species_column), regex("\\ssp\\d+", ignore_case = TRUE))
    )
  
  kept_count <- nrow(df_clean)
  removed_count <- original_count - kept_count
  removed_pct <- round((removed_count / original_count) * 100, 2)
  
  cat("Original rows:", original_count, "\n")
  cat("Rows removed:", removed_count, "\n")
  cat("Rows kept:", kept_count, "\n")
  cat("Percentage removed:", removed_pct, "%\n")
  
  return(df_clean)
}
```

# Function to convert 4 & 6-letter bird alpha codes, and common names, to genus/species and add resolved_name column
```{r}
resolve_bird_names <- function(df, species_column = "resolved_name") {
  # Step 1: Prepare bird reference table
  bird_df <- birdnames::bird_list %>%
    mutate(
      alpha_code_upper = toupper(alpha_code),
      common_clean = tolower(gsub("[^a-z]", "", common_name)),
      scientific_name = str_trim(scientific_name)
    )

  # Step 2: Normalize input species names from the chosen column (usually resolved_name)
  df <- df %>%
    mutate(
      input_upper = toupper(!!sym(species_column)),
      input_clean = tolower(gsub("[^a-z]", "", !!sym(species_column)))
    )

  # Step 3: Attempt alpha code match (e.g., "AMRO" → "Turdus migratorius")
  df <- df %>%
    left_join(
      bird_df %>% select(alpha_code_upper, scientific_name),
      by = c("input_upper" = "alpha_code_upper")
    ) %>%
    mutate(
      bird_resolved = scientific_name  # save first round of matches
    ) %>%
    select(-scientific_name)

  # Step 4: Attempt common name match if no alpha match worked
  df <- df %>%
    left_join(
      bird_df %>% select(common_clean, scientific_name),
      by = c("input_clean" = "common_clean")
    ) %>%
    mutate(
      # Use alpha code match if available, else common name match
      bird_resolved = coalesce(bird_resolved, scientific_name),

      # Only update resolved_name if bird_resolved found
      resolved_name = ifelse(!is.na(bird_resolved), bird_resolved, !!sym(species_column))
    ) %>%
    select(-input_upper, -input_clean, -scientific_name, -bird_resolved)

  return(df)
}
```

# Function to convert 7 & 8-character plant codes to genus/species and add resolved_name column
```{r}
resolve_plant_codes <- function(df, species_column = "resolved_name") {
  # Step 1: Identify likely plant codes (uppercase, 7–8 characters)
  df <- df %>%
    mutate(
      is_plant_code = nchar(!!sym(species_column)) %in% c(7, 8) &
                      grepl("^[A-Z]+$", !!sym(species_column))
    )

  # Step 2: Generate genus/species guesses from the codes
  df <- df %>%
    mutate(
      genus_guess = ifelse(is_plant_code,
                           str_to_title(substr(!!sym(species_column), 1, 4)),
                           NA),
      species_guess = ifelse(is_plant_code,
                             tolower(substr(!!sym(species_column), 5, nchar(!!sym(species_column)))),
                             NA),
      candidate_name = ifelse(is_plant_code,
                              paste(genus_guess, species_guess),
                              NA)
    )

  # Step 3: Resolve guesses using taxize::gnr_resolve
  candidate_names <- unique(na.omit(df$candidate_name))

  resolved <- tryCatch({
    gnr_resolve(candidate_names, best_match_only = TRUE)
  }, error = function(e) {
    warning("gnr_resolve failed: ", e$message)
    return(data.frame(user_supplied_name = character(), matched_name2 = character()))
  })

  # Step 4: Join resolved names back into the data
  if ("matched_name2" %in% names(resolved)) {
    df <- df %>%
      left_join(resolved, by = c("candidate_name" = "user_supplied_name")) %>%
      mutate(
        # Replace resolved_name only if a valid match was found
        resolved_name = ifelse(!is.na(matched_name2), matched_name2, !!sym(species_column))
      )
  } else {
    warning("No matched_name2 column found in resolved names; skipping resolution")
    df$resolved_name <- df[[species_column]]  # Fallback: preserve input
  }

  # Step 5: Clean up helper columns
  df <- df %>%
    select(-is_plant_code, -genus_guess, -species_guess, -candidate_name,
           -matched_name2, -data_source_title)

  return(df)
}
```

# Function to spellcheck valid_name using TNRS and add resolved_name column
```{r}
spellcheck_species_names <- function(df, species_column = "resolved_name") {
  # Step 1: Get unique, non-missing values from the working column (e.g., resolved_name)
  name_vector <- unique(na.omit(df[[species_column]]))

  # Step 2: Resolve names via TNRS, wrapped to avoid crashing on API failure
  resolved <- tryCatch({
    tnrs(name_vector)
  }, error = function(e) {
    warning("TNRS lookup failed: ", e$message)
    return(data.frame(submitted_name = character(), matched_name = character()))
  })

  # Step 3: Ensure result is usable
  if (nrow(resolved) == 0 || !"matched_name" %in% names(resolved)) return(df)

  # Step 4: Normalize names and filter for actual spelling differences
  normalize_name <- function(x) {
    x %>%
      tolower() %>%
      gsub("[^a-z]", "", .)  # Strip all non-letter characters
  }

  resolved <- resolved %>%
    select(submitted_name, matched_name) %>%
    filter(!is.na(matched_name)) %>%
    mutate(
      submitted_norm = normalize_name(submitted_name),
      matched_norm = normalize_name(matched_name)
    ) %>%
    filter(submitted_norm != matched_norm)  # Only keep meaningful spelling differences

  # Step 5: Apply corrections back to the resolved_name column only
  if (nrow(resolved) == 0) return(df)

  df <- df %>%
    left_join(resolved, by = setNames("submitted_name", species_column)) %>%
    mutate(
      !!species_column := ifelse(!is.na(matched_name), matched_name, !!sym(species_column))
    ) %>%
    select(-matched_name, -submitted_norm, -matched_norm)

  return(df)
}
```

# Wrapping function to resolve all species names to genus/species
```{r}

resolve_species_names <- function(df, species_column = "valid_name") {
  cat("Starting species name resolution pipeline...\n")
  
  # Step 0: Create resolved_name column as a copy of the original
  df <- df %>%
    mutate(resolved_name = !!sym(species_column))

  # Step 1: Resolve bird alpha codes and common names
  cat("- Resolving bird names...\n")
  df <- resolve_bird_names(df, species_column = "resolved_name")

  # Step 2: Resolve plant 7–8 character codes
  cat("- Resolving plant codes...\n")
  df <- resolve_plant_codes(df, species_column = "resolved_name")

  # Step 3: Spellcheck using TNRS
  cat("- Spellchecking remaining names...\n")
  df <- spellcheck_species_names(df, species_column = "resolved_name")

  cat("Name resolution complete.\n")
  return(df)
}


```

# Apply filtering and name resolving
```{r}
# Step 1: filter unusable data
filtered_data <- filter_species(biotime_data)

# Step 2: resolve species names
resolved_data <- resolve_species_names(filtered_data, species_column = "valid_name")

# step 3: export the cleaned & resolved dataset
write.csv(as.data.frame(resolved_data), "filtered_resolved_biotimes.csv", row.names = FALSE)

sum(resolved_data$valid_name != resolved_data$resolved_name, na.rm = TRUE)
# Removed 2.13% of rows with all filters applied (255,380 out of 11,989,233)
# head(filtered_data, 50)
```

# Extract unique_genus_species list from the resolved database
```{r}
unique_genus_species <- resolved_data %>%
  filter(!is.na(resolved_name)) %>%
  filter(str_count(resolved_name, "\\s+") >= 1) %>%  # must have at least two words
  mutate(
    GENUS = word(resolved_name, 1),
    SPECIES = word(resolved_name, 2),
    GENUS_SPECIES = paste(GENUS, SPECIES)
  ) %>%
  filter(!is.na(GENUS), !is.na(SPECIES)) %>%
  distinct(GENUS_SPECIES, .keep_all = TRUE) %>%
  arrange(GENUS, SPECIES) %>%
  mutate(NUMBER = row_number()) %>%
  select(NUMBER, GENUS, SPECIES, GENUS_SPECIES)

write.csv(unique_genus_species, "unique_genus_species_biotime.csv", row.names = FALSE)
```












# STAN's PREVIOUS WORK
```{r}
tmp3 = tmp2 %>%select(GENUS_SPECIES) %>%
  distinct() # obtain list of unique species

# create subset to see if R and slurm script works
tmp3_1 = tmp3[1:4000,] %>% # adjust 4000 to smaller values for faster testing
  as.data.frame() %>%
  rename("GENUS_SPECIES" = ".") 

save(tmp3, file = "taxonomies.RData")
save(tmp3_1, file = "taxonomies_1.RData")
```

# in-situ testing of code
## taxize it to add family
```{r}
library(taxize)

taxize_options(ncbi_sleep = 0.8) # you can adjust this time to prevent api limit exceed

tmp_taxized2 <- classification(tmp3_1$GENUS_SPECIES, db = "ncbi",
                              batch_size = 5) # tap on ncbi db to add taxonomic values to species; recommend setting tmp3_1 to just 5-10 rows for fast results
```

```{r}
extract_family <- function(df) {
  if (is.data.frame(df)) {
   family_row <- df[df$rank == "family", ]
   order_row = df[df$rank == "order", ] 
   class_row = df[df$rank == "class",]
   return(paste(family_row$name, order_row$name, class_row$name,
                sep="_"))
   } else {
   return(NA)
   }
}

ltfem_seedling_families2 <- lapply(tmp_taxized2, extract_family)

ltfem_seedling_families2 <- ltfem_seedling_families2[!sapply(ltfem_seedling_families2, is.na)]

ltfem_seedling_families2 = as.data.frame(ltfem_seedling_families2) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "species")
# 
# ltfem_seedling_families$species =  str_replace_all(ltfem_seedling_families$species,"\\."," ")
# 
# ltfem_seedling_families = ltfem_seedling_families %>%
#   separate(V1, into = c("family", "order"), sep = "_")
```


# extract class and other taxonomic values
this chunk below is for obtaining the results from the slurm job
```{r}

load("tmp_foc.RData")
load("tmp_taxized_slurm.RData")

# run the code below only if the taxize.R script did not do the following post-processing
tmp_foc <- tmp_foc[sapply(tmp_foc, length) > 0]

tmp = as.data.frame(tmp_foc) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "species")

tmp$species =  str_replace_all(tmp$species,"\\."," ")

tmp = tmp %>%
  separate(V1, into = c("family", "order", "class"), sep = "_")

# export tmp to view results
```



